---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# R2spa

<!-- badges: start -->
[![R-CMD-check](https://github.com/mmm-lab-um/R2spa/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mmm-lab-um/R2spa/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`R2spa` is a free and open-source R package that performs two-stage path analysis (2S-PA). With 2S-PA, researchers can perform path analysis by first obtaining factor scores and then adjusting for measurement errors using estimates of observation-specific reliability or standard error of those factor scores. As a viable alternative to SEM, 2S-PA has been shown to give equally-good estimates as SEM in relatively simple models and large sample sizes, as well as to give more accurate parameter estimates, has better control of Type I error rates, and has substantially less convergence problems in more complex models or small sample sizes.

## Installation

This package is still in developmental stage and can be installed on GitHub with:

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_github("mmm-lab-um/R2spa")
```

## Documentation

The package website, <https://mmm-lab-um.github.io/R2spa/>, hosts the full
reference and a series of worked articles. Start with the
[Two-Stage Path Analysis (2S-PA) Model Examples](https://mmm-lab-um.github.io/R2spa/articles/R2spa.html)
getting-started article.

## Example

The canonical workflow scores each latent from one multi-factor measurement
model (`get_fs(..., local = TRUE)`), then fits the structural path with
`tspa()`, which reads each score's standard error from the `fs_<name>_se`
columns automatically (no hard-coded `se_fs` needed):

```{r message=FALSE}
library(lavaan)
library(R2spa)

# One multi-factor measurement model
meas_model <- '
  ind60 =~ x1 + x2 + x3
  dem60 =~ y1 + y2 + y3 + y4
  dem65 =~ y5 + y6 + y7 + y8
'
```

```{r}
# Stage 1: score each latent from its own measurement model (local = TRUE)
fs_dat <- get_fs(PoliticalDemocracy, model = meas_model, local = TRUE)
# get_fs() gives a data frame of factor scores and their standard errors
head(fs_dat)
```

```{r}
# Stage 2: fit the structural path; se_fs is read from the fs_*_se columns
tspa_fit <- tspa(
  model = "dem60 ~ ind60
          dem65 ~ ind60 + dem60",
  data = fs_dat
)
```

Because the latent constructs have no intrinsic scale, the **standardized**
coefficients are usually the parameters of substantive interest (see the
*Two-Stage Path Analysis* vignette, `vignette("R2spa")`, for the reasoning and
references):

```{r}
standardizedSolution(tspa_fit) |>
  subset(op %in% c("~", "~~") & !grepl("^fs_", lhs))
```

> **Optional dependencies.** The exact (non-pooled) stage-2 route,
> `tspa_mx_model()`, requires `OpenMx`, and the item-response-theory
> (`mirt`) scoring path requires `mirt` (both `Suggests`); see the
> *2S-PA with OpenMx and IRT (mirt)* vignette,
> `vignette("tspa-vignette-mx")`.

This package is based upon work supported by the National Science Foundation under Grant No. 2141790.

<!-- `devtools::build_readme()` -->

