---
title: "Corrected Standard Errors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Corrected Standard Errors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
  - \DeclareMathOperator{\Var}{\mathrm{Var}}
  - \DeclareMathOperator{\E}{\mathrm{E}}
  - \newcommand{\bv}[1]{\boldsymbol{\mathbf{#1}}}  %APA-consistent bold
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows how to obtain Taylor-series corrected standard errors in 2S-PA, accounting for the uncertainty in the weights for obtaining the factor scores. It is described in [this paper](https://doi.org/10.1007/s00181-020-01942-z).

The second-stage estimation treats as known the loading matrix and the error covariance matrix of the factor scores as indicators of the true latent variables. However, those are estimates based on the first-stage measurement model. When the sample size is large and/or the reliability of the factor scores is high, the uncertainty in the estimated loading and error covariance matrix is generally negligible. Otherwise, a first-order correction on the standard errors of the second-stage estimation is possible, as illustrated below.

## First-order correction for SE

$$\hat V_{\gamma, c} = \hat V_{\gamma} + \boldsymbol{J}_\boldsymbol{\gamma}(\hat{\boldsymbol{\theta}}) \hat V_{\theta} \boldsymbol{J}_\boldsymbol{\gamma}(\hat{\boldsymbol{\theta}})^\top,$$

where $\boldsymbol{J}_\boldsymbol{\gamma}$ is the Jacobian matrix of $\hat{\boldsymbol{\gamma}}$ with respect to $\boldsymbol{\theta}$, or

$$\hat V_{\gamma, c} = \hat V_{\gamma} + (\boldsymbol{H}_\gamma)^{-1} \left(\frac{\partial^2 \ell}{\partial \theta \partial \gamma^\top}\right) \hat V_{\theta} \left(\frac{\partial^2 \ell}{\partial \theta \partial \gamma^\top}\right)^\top (\boldsymbol{H}_\gamma)^{-1},$$

where $V_{\gamma}$ is the naive covariance matrix of the structural parameter estimates $\hat{\boldsymbol{\gamma}}$ assuming the measurement model parameters $\boldsymbol{\theta}$ (the score loadings and error variances) are known, $\boldsymbol{H}_\gamma$ is the Hessian matrix of the log-likelihood $\ell$ with respect to $\hat{\boldsymbol{\gamma}}$, and $V_{\theta}$ can be obtained in the first-stage measurement model analysis.

```{r load-pkg, message = FALSE}
library(lavaan)
library(R2spa)
library(numDeriv)
library(boot)
```

This example is based on the `PoliticalDemocracy` dataset, from the lavaan tutorial (<https://lavaan.ugent.be/tutorial/sem.html>), also used as an example in Bollen (1989).

## Separate Measurement Models

```{r}
model <- '
  # latent variable definitions
     ind60 =~ x1 + x2 + x3
     dem60 =~ y1 + y2 + y3 + y4

  # regressions
    dem60 ~ ind60
'
```

```{r}
cfa_ind60 <- cfa("ind60 =~ x1 + x2 + x3", data = PoliticalDemocracy)
# Regression factor scores
fs1 <- get_fs(cfa_ind60, vfsLT = TRUE, format = "list")
# Delta method variance of (loading, error variance)
vldev1 <- attr(fs1, which = "vfsLT")
cfa_dem60 <- cfa("dem60 =~ y1 + y2 + y3 + y4",
                 data = PoliticalDemocracy)
# Regression factor scores
fs2 <- get_fs(cfa_dem60, vfsLT = TRUE, format = "list")
# Delta method variance of (loading, error variance)
vldev2 <- attr(fs2, which = "vfsLT")
fs_dat <- data.frame(
  fs_ind60 = fs1$fs_ind60,
  fs_dem60 = fs2$fs_dem60
)
# Combine sampling variance of loading and error variance
# Note: loadings first, then error variance
vldev <- block_diag(vldev1, vldev2)[c(1, 3, 2, 4), c(1, 3, 2, 4)]
```

```{r}
# 2S-PA
# Assemble loadings
ld <- block_diag(attr(fs1, which = "fsL"),
                 attr(fs2, which = "fsL"))
ev <- block_diag(attr(fs1, which = "fsT"),
                 attr(fs2, which = "fsT"))
tspa_fit <- tspa(model = "dem60 ~ ind60",
                 data = fs_dat,
                 fsL = ld,
                 fsT = ev)
# Unadjusted covariance
vcov(tspa_fit)
# Adjusted covariance matrix
(vc_cor <- vcov_corrected(tspa_fit, vfsLT = vldev, which_free = c(1, 4, 5, 7)))
# Corrected standard errors
sqrt(diag(vc_cor))
```

The `engine` argument of `vcov_corrected()` (and of `tspa(corrected_se = TRUE)`) selects how the Jacobian is evaluated. The default `"fd"` uses central finite differences (one stage-2 refit on each side of each free element); `"analytic"` evaluates the same Jacobian refit-free and deterministically for the (approximately) saturated single-group case, transparently falling back to `"fd"` otherwise (multigroup, or a restricted structural model). The two agree to the finite-difference noise floor whenever `"analytic"` applies, and the analytic result is bit-reproducible.

Standardized solution

The corrected standard error of the `dem60 ~ ind60` coefficient, derived by hand as the standard error of the standardized coefficient plus a first-order delta-method term:

```{r}
keep_dem_ind <- function(std) {
  std[std$lhs == "dem60" & std$rhs == "ind60" & std$op == "~", , drop = FALSE]
}
tspa_est_std <- function(ld_ev) {
  ld1 <- ld
  ev1 <- ev
  ld1[c(1, 4)] <- ld_ev[1:2]
  ev1[c(1, 4)] <- ld_ev[3:4]
  tfit <- tspa(model = "dem60 ~ ind60",
               data = fs_dat,
               fsL = ld1,
               fsT = ev1)
  keep_dem_ind(standardizedSolution(tfit))$est.std
}
tspa_est_std(c(ld[c(1, 4)], ev[c(1, 4)]))
# Jacobian of the standardized coefficient w.r.t. (loadings, error variances)
jac_std <- numDeriv::jacobian(tspa_est_std,
                              x = c(ld[c(1, 4)], ev[c(1, 4)]))
ss0 <- keep_dem_ind(standardizedSolution(tspa_fit))
sqrt(ss0$se^2 + jac_std %*% vldev %*% t(jac_std))
```

The same value comes from the one-call option. Because the two factors were fitted in separate models, only the two loadings and the two error variances are free (positions 1, 4, 5, 7), so we pass the matching 4 x 4 submatrix of `vldev` together with `which_free`:

```{r}
tspa_fit_corr <- tspa(model = "dem60 ~ ind60",
                      data = fs_dat,
                      fsL = ld,
                      fsT = ev,
                      vfsLT = vldev,
                      which_free = c(1, 4, 5, 7),
                      corrected_se = TRUE)
sc <- keep_dem_ind(standardizedSolution(tspa_fit_corr))
sc$est.std              # unchanged point estimate
sc$se                   # corrected standard SE (matches the hand-derived value)
attr(tspa_fit_corr, "tspa_corrected")   # TRUE
```

Compare to joint structural and measurement model (unstandardized solution):

```{r}
sem_fit <- sem(model, data = PoliticalDemocracy)
# Larger standard error
(vc_j <- 
  vcov(sem_fit)[c("dem60~ind60", "ind60~~ind60", "dem60~~dem60"),
                c("dem60~ind60", "ind60~~ind60", "dem60~~dem60")])
sqrt(diag(vc_j))
```

Bootstrap Standard Errors

```{r, eval = FALSE, echo = TRUE}
run_tspa <- function(df, inds) {
  # local = TRUE: score each latent from its own (separate) measurement model
  fs_dat <- get_fs(object = df[inds, ],
                   model = "ind60 =~ x1 + x2 + x3
                            dem60 =~ y1 + y2 + y3 + y4",
                   local = TRUE,
                   se = "none", test = "none")
  # tspa() reads fsL/fsT/fsb and the per-score SEs from the get_fs() result
  tspa_fit <- tspa(model = "dem60 ~ ind60",
                   data = fs_dat,
                   test = "none")
  coef(tspa_fit)
}
boo <- boot(PoliticalDemocracy, statistic = run_tspa, R = 1999)
```

```{r, eval = FALSE, include = FALSE}
saveRDS(boo, file = "boo_separate.RDS")
```

```{r, include = FALSE}
boo <- readRDS("boo_separate.RDS")
```

```{r}
# Use MAD to downweigh outlying replications
boo$t |>
    apply(MARGIN = 2, FUN = mad) |>
    setNames(c("dem60~ind60", "ind60~~ind60", "dem60~~dem60"))
```

The standard errors seem to diverge slightly among methods. SE(`dem60~ind60`) with bootstrap is the lowest. SE(`ind60~~ind60`) with the joint model is particularly higher than the other two. SE(`dem60~~dem60`) is the lowest with the corrected 2S-PA. It should be pointed out that the joint model and 2S-PA are different estimators.

```{r, eval = FALSE}
run_sem <- function(df, inds) {
  sem_fit <- sem(model, data = df[inds, ], se = "none", test = "none")
  coef(sem_fit)
}
boo <- boot(PoliticalDemocracy, statistic = run_sem, R = 999)
```

## Joint Measurement Model

```{r}
cfa_joint <- cfa("ind60 =~ x1 + x2 + x3
                  dem60 =~ y1 + y2 + y3 + y4",
                 data = PoliticalDemocracy)
# Factor score
fs_joint <- get_fs_lavaan(cfa_joint, vfsLT = TRUE)
# Delta method variance
vldev_joint <- attr(fs_joint, which = "vfsLT")
```

```{r}
tspa_fit2 <- tspa(model = "dem60 ~ ind60",
                  data = fs_joint,
                  fsT = attr(fs_joint, "fsT"),
                  fsL = attr(fs_joint, "fsL"))
# Unadjusted covariance
vcov(tspa_fit2)
# Adjusted covariance
(vc2_cor <- vcov_corrected(tspa_fit2, vfsLT = vldev_joint))
# Corrected standard errors
sqrt(diag(vc2_cor))
```

Bootstrap Standard Errors

```{r, eval = FALSE}
run_tspa2 <- function(df, inds) {
  fs_joint <- get_fs(df[inds, ],
                     model = "ind60 =~ x1 + x2 + x3
                              dem60 =~ y1 + y2 + y3 + y4",
                     se = "none", test = "none")
  # tspa() reads fsL/fsT/fsb and the per-score SEs from the get_fs() result
  tspa_fit2 <- tspa(model = "dem60 ~ ind60",
                    data = fs_joint,
                    test = "none")
  coef(tspa_fit2)
}
boo2 <- boot(PoliticalDemocracy, statistic = run_tspa2, R = 1999)
# run_tspa2b <- function(df, inds) {
#   fsb_joint <- get_fs(df[inds, ],
#                       model = "ind60 =~ x1 + x2 + x3
#                                dem60 =~ y1 + y2 + y3 + y4",
#                       se = "none", test = "none", method = "Bartlett")
#   tspa_fit2b <- tspa(model = "dem60 ~ ind60",
#                      data = fsb_joint,
#                      fsT = attr(fsb_joint, "fsT"),
#                      fsL = diag(2),
#                      test = "none")
#   coef(tspa_fit2b)
# }
# boo2b <- boot(PoliticalDemocracy, statistic = run_tspa2b, R = 4999)
# run_sem2 <- function(df, inds) {
#   sem_fit <- sem(model, data = df[inds, ])
#   coef(sem_fit)[c(6, 14, 15)]
# }
# boo2j <- boot(PoliticalDemocracy, statistic = run_sem2, R = 4999)
```

```{r, eval = FALSE, include = FALSE}
saveRDS(boo2, file = "boo_joint.RDS")
```

```{r, include = FALSE}
boo2 <- readRDS("boo_joint.RDS")
```

```{r}
# Use MAD to downweigh outlying replications
boo2$t |>
    apply(MARGIN = 2, FUN = mad) |>
    setNames(c("dem60~ind60", "ind60~~ind60", "dem60~~dem60"))
```

### With Bartlett's Method

```{r}
# Factor score
fsb_joint <- get_fs(PoliticalDemocracy,
                    model = "ind60 =~ x1 + x2 + x3
                             dem60 =~ y1 + y2 + y3 + y4",
                    method = "Bartlett",
                    vfsLT = TRUE)
vldevb_joint <- attr(fsb_joint, which = "vfsLT")
```

```{r}
tspa_fit2b <- tspa(model = "dem60 ~ ind60",
                   data = fsb_joint,
                   fsT = attr(fsb_joint, "fsT"),
                   fsL = diag(2) |>
                       `dimnames<-`(list(c("fs_ind60", "fs_dem60"),
                                         c("ind60", "dem60"))))
# Unadjusted covariance matrix
vcov(tspa_fit2b)
# Adjusted covariance matrix
(vc2b_cor <- vcov_corrected(
    tspa_fit2b,
    # With Bartlett scoring the loadings are fixed, so propagate the error
    # variances only (positions 5 and 7)
    vfsLT = vldevb_joint[c(5, 7), c(5, 7)],
    which_free = c(5, 7)))
# Corrected standard errors
sqrt(diag(vc2b_cor))
```

## Multiple Groups

```{r}
# Multigroup, three-factor example
mod <- "
  # latent variables
    visual =~ x1 + x2 + x3
    textual =~ x4 + x5 + x6
    speed =~ x7 + x8 + x9
"
# Factor scores based on partial invariance
fit <- cfa(mod,
           data = HolzingerSwineford1939,
           std.lv = TRUE,
           group = "school",
           group.equal = c("loadings", "intercepts"),
           group.partial = c("visual=~x2", "x7~1"))
fs_dat <- get_fs_lavaan(fit, vfsLT = TRUE)
vldev <- attr(fs_dat, which = "vfsLT")
```

```{r}
tspa_fit <- tspa(model = "visual ~~ textual + speed
                          textual ~~ speed",
                 data = fs_dat,
                 group = "school",
                 fsL = attr(fs_dat, which = "fsL"),
                 fsT = attr(fs_dat, which = "fsT"))
# Unadjusted covariance of the structural parameters (the 12 factor
# variances/covariances, 6 per group); the 6 factor-score intercepts (fs_*~1)
# are omitted.
keep <- grepl("~~", rownames(vcov(tspa_fit)))
vcov(tspa_fit)[keep, keep]
# Adjusted covariance
vcov_corrected(tspa_fit, vfsLT = vldev)[keep, keep]
```

The correction can be requested in place on `tspa()`, so `vcov()` (and hence the SEs derived from it) and `standardizedSolution()` on the returned fit report the corrected standard errors (the point estimates are unchanged):

```{r}
tspa_fit_corr <- tspa(model = "visual ~~ textual + speed
                               textual ~~ speed",
                      data = fs_dat,
                      group = "school",
                      fsL = attr(fs_dat, which = "fsL"),
                      fsT = attr(fs_dat, which = "fsT"),
                      vfsLT = vldev,
                      corrected_se = TRUE)
attr(tspa_fit_corr, "tspa_corrected")   # TRUE
round(cbind(unadjusted = sqrt(diag(vcov(tspa_fit))),
            corrected    = sqrt(diag(vcov(tspa_fit_corr)))), 4)
```

### Corrected grand-standardized solution (multigroup)

`grandStandardizedSolution()` pools across groups (grand mean / grand SD) rather than standardizing each group separately as `lavaan::standardizedSolution()` does. Feeding it the corrected fit's covariance yields the corrected grand-standardized SEs; we use a structural path model so there is a regression to standardize:

```{r}
path_mg <- "textual ~ visual + speed
            visual  ~ speed"
tspa_path  <- tspa(path_mg, data = fs_dat, group = "school",
                   fsL = attr(fs_dat, "fsL"), fsT = attr(fs_dat, "fsT"))
tspa_path_corr <- tspa(path_mg, data = fs_dat, group = "school",
                       fsL = attr(fs_dat, "fsL"), fsT = attr(fs_dat, "fsT"),
                       vfsLT = vldev, corrected_se = TRUE)
gr0 <- grandStandardizedSolution(tspa_path)       # uncorrected
gr1 <- grandStandardizedSolution(tspa_path_corr)  # corrected (SE only)
keep <- gr1$op == "~"
round(cbind(grand_std = gr1$est.std[keep],
            se_plain  = gr0$se[keep],
            se_corr   = gr1$se[keep]), 4)
# SE-only: identical point estimates, corrected SEs >= uncorrected
all(gr1$est.std[keep] == gr0$est.std[keep])
all(gr1$se[keep] >= gr0$se[keep])
```


```{r}
keep_dem_ind <- function(std) {
  std[std$lhs %in% c("visual", "textual", "speed") &
    std$rhs %in% c("visual", "textual", "speed") & 
      std$lhs != std$rhs, , drop = FALSE]
}
tspa_est_std <- function(ld_ev) {
  fsL1 <- attr(fs_dat, which = "fsL")
  fsT1 <- attr(fs_dat, which = "fsT")
  fsL1[[1]][] <- ld_ev[1:9]
  fsL1[[2]][] <- ld_ev[10:18]
  fsT1[[1]][] <- ld_ev[c(19:21, 20, 22:23, 21, 23, 24)]
  fsT1[[2]][] <- ld_ev[c(25:27, 26, 28:29, 27, 29, 30)]
  tfit <- tspa(model = "visual ~~ textual + speed
                        textual ~~ speed",
               data = fs_dat,
               group = "school",
               fsL = fsL1,
               fsT = fsT1) 
  keep_dem_ind(standardizedSolution(tfit))$est.std
}
ld_ev0 <- c(attr(fs_dat, which = "fsL")[[1]],
            attr(fs_dat, which = "fsL")[[2]],
            attr(fs_dat, which = "fsT")[[1]][c(1:3, 5:6, 9)],
            attr(fs_dat, which = "fsT")[[2]][c(1:3, 5:6, 9)])
tspa_est_std(ld_ev0)
jac <- numDeriv::jacobian(tspa_est_std, x = ld_ev0)
v0 <- lavInspect(tspa_fit, "vcov.std.lv")[c(1:3, 10:12), c(1:3, 10:12)]
sqrt(diag(jac %*% vldev %*% t(jac)) / diag(v0))

```