---
title: "EFA Scores"
author: "Mark Lai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EFA Scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how to use the `compute_fscore()` function to calculate factor scores based on exploratory factor analysis, and compare the results to those calculated by the `psych::factor.scores()` function.

```{r}
library(psych)
data(bfi, package = "psych")
library(lavaan)
```

## Example: Big Five Inventory

```{r}
# Covariance (with FIML)
corr_bfi <- lavCor(bfi[1:25], missing = "fiml")
# EFA (Target rotation)
target_mat_bfi <- matrix(0, nrow = 25, ncol = 5)
target_mat_bfi[1:5, 1] <- NA
target_mat_bfi[6:10, 2] <- NA
target_mat_bfi[11:15, 3] <- NA
target_mat_bfi[16:20, 4] <- NA
target_mat_bfi[21:25, 5] <- NA
fa_target_bfi <- psych::fa(
    corr_bfi, n.obs = 2436,
    nfactors = 5,
    rotate = "targetQ", Target = target_mat_bfi,
    scores = "Bartlett")
# Factor correlations
fa_target_bfi$Phi |>
    (`[`)(paste0("MR", 1:5), paste0("MR", 1:5)) |>
    knitr::kable(digits = 2, caption = "EFA Factor Correlation")
```

```{r}
# Correlation with sum scores
bfi |>
    transform(A = (7 - A1) + A2 + A3 + A4 + A5,
           C = C1 + C2 + C3 + (7 - C4) + (7 - C5),
           E = (7 - E1) + (7 - E2) + E3 + E4 + E5,
           N = N1 + N2 + N3 + N4 + N5,
           O = O1 + (7 - O2) + O3 + O4 + (7 - O5)) |>
    subset(select = A:O) |>
    cor(use = "complete") |>
    knitr::kable(digits = 2, caption = "Sum scores")
```

Hand calculate the Bartlett scores using weights

```{r}
# Bartlett score for first person
bscores <-
    psych::factor.scores(bfi[, 1:25], f = fa_target_bfi,
                         method = "Bartlett")
fa_target_bfi$weights
# Calculation by hand
y1 <- scale(bfi[, 1:25])[1, ]  # z-score
crossprod(fa_target_bfi$weights, as.matrix(y1))
# Compare to results from psych::fa()
bscores$scores[1, ]
```

```{r}
# Covariance of Bartlett scores
cov(bscores$scores, use = "complete") |>
    (`[`)(paste0("MR", 1:5), paste0("MR", 1:5)) |>
    knitr::kable(digits = 2, caption = "With Bartlett scores")
```

### Using `compute_fscore()` and perform a two-stage analysis

Use `R2spa::compute_fscore()`

```{r}
# Obtain error covariances
yc <- scale(bfi[, 1:25])
yc <- yc[complete.cases(yc), ]
lam <- fa_target_bfi$loadings
colnames(lam) <- c("N", "E", "C", "A", "O")
phi <- fa_target_bfi$Phi
th <- diag(fa_target_bfi$uniquenesses)
# # scoring weights
# a <- solve(crossprod(lam, solve(th, lam)), t(solve(th, lam)))
# ecov_fs <- a %*% th %*% t(a)
# dimnames(ecov_fs) <- rep(list(c("N", "E", "C", "A", "O")), 2)
# Two-stage analysis
library(R2spa)
bfi_fs <- compute_fscore(yc, lambda = lam, theta = th,
                         method = "Bartlett", center_y = FALSE,
                         fs_matrices = TRUE)
head(bfi_fs)
# Scoring matrix
attr(bfi_fs, which = "scoring_matrix")
# Error covariance
attr(bfi_fs, which = "fsT")
```

Recover factor covariances with 2S-PA

```{r}
ts_fit <- tspa("",
               data = data.frame(bscores$scores) |>
                   setNames(c("fs_N", "fs_E", "fs_C", "fs_A", "fs_O")),
               fsT = attr(bfi_fs, which = "fsT"),
               fsL = diag(5) |>
                   `dimnames<-`(list(c("fs_N", "fs_E", "fs_C", "fs_A", "fs_O"),
                                     c("N", "E", "C", "A", "O"))))
lavInspect(ts_fit, what = "cor.lv") |>
    (`[`)(c("A", "C", "E", "N", "O"), c("A", "C", "E", "N", "O")) |>
    knitr::kable(digits = 2, caption = "With Bartlett scores and 2S-PA")
```


