---
title: "2S-PA with Missing Data"
author: "Mark Lai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{missing-data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

With missing data, two things change in two-stage path analysis. Stage 1
must be fitted with a missing-data method (`missing = "fiml"` in `lavaan`),
and each case is then scored on its *own* observed indicators, so the score
standard errors, implied loadings, and error variances are **per-row**
quantities rather than one pooled matrix. And stage 2 has to use those
per-row values: `tspa()` (the `lavaan` backend) reduces the per-pattern
values to a single representative matrix (`reduce = `), whereas
`tspa_mx_model()` (the `OpenMx` backend) carries the per-row values through
definition variables, and its raw-data FIML evaluation can also handle rows
whose scores are dropped as `NA`. The stage-2 fits in this vignette therefore
use `tspa_mx_model()`.

For illustration, we create two missing data patterns on
`PoliticalDemocracy`: (1) MCAR missingness in a single item (`x1`, about 30%
of cases); (2) block missingness, where the last 10 cases are missing all
four `dem65` indicators — a factor with no observed indicator at all.

## Single-Factor Model

```{r}
set.seed(2225)
library(lavaan)
library(R2spa)
library(OpenMx)
data(PoliticalDemocracy)
pd2 <- PoliticalDemocracy
# Add MCAR missing data to x1 (about 30% of the 75 cases)
pd2[!rbinom(75, size = 1, prob = 0.7), "x1"] <- NA
```

Factor scores with `lavaan::lavPredict()`:

```{r}
fit <- cfa("ind60 =~ x1 + x2 + x3", data = pd2, missing = "fiml")
fs_lavaan <- lavPredict(fit, method = "Bartlett", se = TRUE, fsm = TRUE)
# From R2spa
fs <- R2spa::get_fs(fit, method = "Bartlett")
# One SE per missing-data pattern
round(table(pattern = attr(fs, "fs_pattern")[[1]]$label,
            se = fs$fs_ind60_se), 4)
# Compare factor scores
plot(fs_lavaan, fs$fs_ind60)
abline(a = 0, b = 1)
```

Because each case is scored conditionally on its own observed indicators,
`get_fs()` returns per-row standard errors and error variances: the 48
complete cases (pattern `"x1+x2+x3"`) get SE 0.150, and the 27 cases observed
only on `x2` and `x3` get SE 0.179. The `fsT`/`fsL`/`fsb`/`scoring_matrix`
attributes are correspondingly *lists* with one entry per pattern, and the
`fs_pattern` attribute maps each row to its pattern. The plot confirms that
the `R2spa` scores reproduce `lavPredict()` exactly.

### Multiple group

```{r}
mg_fit <- cfa("ind60 =~ x1 + x2 + x3",
  data = cbind(pd2, group = rep(1:2, c(40, 35))),
  missing = "fiml",
  group = "group"
)
mg_fs <- R2spa::get_fs(mg_fit)
head(mg_fs, 10)
# Both groups contain the same two patterns
lapply(attr(mg_fs, "fs_pattern"), function(p) table(p$label))
```

The multigroup result is a single data frame with a `group` column, and each
group's `fsT`/`fsL`/`fsb`/`scoring_matrix` attribute is itself a per-pattern
list.

## Multiple-Factor Model

The measurement model constrains the two factors' loadings to be equal
(labels `a`–`d`) and adds covariances between the corresponding indicators:

```{r}
# Make last 10 cases completely missing on y5-y8 (all dem65 indicators)
pd2[66:75, c("y5", "y6", "y7", "y8")] <- NA
fit2 <- cfa("
  dem60 =~ a * y1 + b * y2 + c * y3 + d * y4
  dem65 =~ a * y5 + b * y6 + c * y7 + d * y8
  y1 ~~ y5
  y2 ~~ y6
  y3 ~~ y7
  y4 ~~ y8
", data = pd2, missing = "fiml")
```

This block-missing pattern is the hard case: for the 10 affected rows,
`dem65` has no observed indicator at all.

### Regression scores

With regression (EBM) scores, the `dem65` score is still *defined* for those
rows — regression scoring uses the cross-factor covariances, so `dem65` is
imputed from the `dem60` indicators. But the per-row measurement information
is degenerate: the implied loading of the score on `dem65` is 0, i.e. the
score carries no information about `dem65`, so measurement-error correction
is impossible. We therefore drop those `dem65` scores; OpenMx's raw-data FIML
then fits those rows from the structural model alone.

```{r}
# Obtain tidy-ed factor scores data
fs_dat <- R2spa::augment_lav_predict(fit2)
# Per-row definition-variable matrices: each cell names the `fs_dat` column
# carrying that case's loadings / error variances (the exact, per-row
# correction that `tspa()`'s `reduce = ` pooling only approximates).
cross_load <- matrix(c(
  "dem60_by_fs_dem60", "dem60_by_fs_dem65",
  "dem65_by_fs_dem60", "dem65_by_fs_dem65"
), nrow = 2) |>
  `dimnames<-`(list(c("fs_dem60", "fs_dem65"), c("dem60", "dem65")))
err_cov <- matrix(c(
  "ev_fs_dem60", "ecov_fs_dem60_fs_dem65",
  "ecov_fs_dem60_fs_dem65", "ev_fs_dem65"
), nrow = 2) |>
  `dimnames<-`(rep(list(c("fs_dem60", "fs_dem65")), 2))
# Drop the non-correctable dem65 scores of the 10 all-missing rows;
# OpenMx's raw-data FIML then fits those rows as usual.
fs_dat[66:75, "fs_dem65"] <- NA
tspa_mx <- tspa_mx_model(
  "dem65 ~ dem60; dem65 + dem60 ~ 1",
  data = fs_dat,
  fsL = cross_load, fsT = err_cov,
  fsb = c(fs_dem60 = "int_fs_dem60", fs_dem65 = "int_fs_dem65")
)
# Run OpenMx
tspa_mx_fit <- mxRun(tspa_mx)
# Summarize the results
summary(tspa_mx_fit)
```

The per-row measurement values reach OpenMx as *definition variables*: `fsL`
and `fsT` are character matrices whose cells name the `fs_dat` column holding
that case's value, so the stage-2 correction is per-row rather than pooled.
`fsb` does the same for the per-row measurement intercepts (the `int_*`
columns; a FIML fit estimates a mean structure, so the intercepts vary with
the missing-data pattern).

### Bartlett scores

Bartlett scoring does not use the cross-factor covariances, so for the 10
rows with no observed `dem65` indicator the score is `NA` by convention — in
contrast to the EBM case above, where the score was defined but
non-correctable and had to be dropped.

```{r}
# Obtain tidy-ed factor scores data (Bartlett weights)
fsb_dat <- augment_lav_predict(fit2, method = "Bartlett")
# fsb_dat[66:75, "fs_dem65"] is already NA (no observed indicator);
# stated explicitly to mirror the EBM case
fsb_dat[66:75, "fs_dem65"] <- NA
tspab_mx <- tspa_mx_model(
  "dem65 ~ dem60; dem65 + dem60 ~ 1",
  data = fsb_dat,
  fsL = cross_load, fsT = err_cov,
  fsb = c(fs_dem60 = "int_fs_dem60", fs_dem65 = "int_fs_dem65")
)
# Run OpenMx
tspab_mx_fit <- mxRun(tspab_mx)
# Summarize the results
summary(tspab_mx_fit)
```

### Compared to Joint Model

The gold-standard reference is the joint model, which fits the measurement
and structural parts simultaneously under FIML:

```{r}
jfit <- sem("
  dem60 =~ a * y1 + b * y2 + c * y3 + d * y4
  dem65 =~ a * y5 + b * y6 + c * y7 + d * y8
  y1 ~~ y5
  y2 ~~ y6
  y3 ~~ y7
  y4 ~~ y8
  dem65 ~ dem60
", data = pd2, missing = "fiml")
summary(jfit)
```

```{r}
# The three estimates of the dem65 ~ dem60 path:
c(EBM.mx = coef(tspa_mx_fit)["m1.A[4,3]"],
  Bartlett.mx = coef(tspab_mx_fit)["m1.A[4,3]"],
  joint = coef(jfit)["dem65~dem60"])
```

Both 2S-PA fits (regression and Bartlett scores) recover the `dem65 ~ dem60`
path at 0.913, identical to the joint-model estimate to four decimal places.

### Multiple group

The same pipeline applies to multigroup fits: `augment_lav_predict()` returns
a list with one tidy-ed data frame per group, each carrying that group's
per-row values. All 10 block-missing rows fall in group 2, so only group 2
contains the pattern with no observed `dem65` indicator. (`lavaan` warns that
a single label used in a multigroup fit imposes cross-group equality — that
is intentional here: `a`–`d` are constrained equal in both groups.)

```{r}
mg_fit2 <- cfa("
  dem60 =~ a * y1 + b * y2 + c * y3 + d * y4
  dem65 =~ a * y5 + b * y6 + c * y7 + d * y8
  y1 ~~ y5
  y2 ~~ y6
  y3 ~~ y7
  y4 ~~ y8
",
  data = cbind(pd2, group = rep(1:2, c(40, 35))),
  missing = "fiml", group = "group"
)
mg_out <- get_fs(mg_fit2)
head(mg_out)
```
