---
title: "2S-PA with Random Effects"
author: "Mark Lai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2S-PA with Random Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette, we show how to use 2S-PA for estimating associations for cluster-level random effects. We use simulated data from a parallel-process growth model, where the two processes have different time scales. This is a pretty challenging model to estimate using standard structural equation models (as time scales are different) or multilevel models (as it is multivariate).

```{r, message = FALSE}
library(R2spa)
library(lme4)
library(OpenMx)
```

The EB (empirical-Bayes) random effects from `lmer()` are per-cluster, so the
per-cluster loadings and error covariances are supplied to `tspa_mx_model()` as
*definition variables* (column names). Each latent is given a name distinct
from its score column (e.g. latent `u0` is measured by the score column
`u0_eb`). A small helper pulls the implied latent covariance back out of the
fitted RAM model by name:

```{r}
lat_cov <- function(fit, lat) {
  v <- c(fit$manifestVars, fit$latentVars)
  idx <- match(lat, v)
  co <- coef(fit)
  out <- matrix(NA_real_, length(lat), length(lat), dimnames = list(lat, lat))
  for (i in seq_along(lat)) for (j in seq_along(lat))
    out[i, j] <- co[sprintf("m1.S[%d,%d]", min(idx[i], idx[j]), max(idx[i], idx[j]))]
  out
}
```

## Simulate bivariate growth data

- Process 1: 5 time points
- Process 2: 20 time points

```{r}
set.seed(1957)
latent_cor <- matrix(
    c(1, 0, 0.5, 0.3,
      0, 1, -0.1, -0.3,
      0.5, -0.1, 1, 0.2,
      0.3, -0.3, 0.2, 1),
    nrow = 4
)
latent_sd <- c(1, 0.5, 1.5, 0.6)
latent_cov <- diag(latent_sd) %*% latent_cor %*% diag(latent_sd)
latent_mean <- c(-1, 0.5, 0, -0.3)
# Simulate person-specific growth parameters
n_clus <- 100
eta <- MASS::mvrnorm(n_clus, mu = latent_mean, Sigma = latent_cov,
                     empirical = TRUE)
# Time variables (different time scale for the two processes)
n_time1 <- 5
n_time2 <- 20
# Assume complete data for now
time1 <- rep(seq_len(n_time1) - 1, n_clus)
time2 <- rep(seq_len(n_time2) - 1, n_clus)
# Simulate process 1
clus_id1 <- rep(seq_len(n_clus), each = n_time1)
sigma1 <- 1  # error sd
# Simulate error
e1 <- rnorm(clus_id1)
e1 <- e1 - mean(e1)
e1 <- e1 / sd(e1) * sigma1
y1 <- eta[clus_id1, 1] + time1 * eta[clus_id1, 2] + e1
# Simulate process 2
clus_id2 <- rep(seq_len(n_clus), each = n_time2)
sigma2 <- 2  # error sd
# Simulate error
e2 <- rnorm(clus_id1)
e2 <- e2 - mean(e2)
e2 <- e2 / sd(e2) * sigma2
y2 <- eta[clus_id2, 3] + time2 * eta[clus_id2, 4] + e2
```

```{r}
# Empirical Bayes
m1 <- lmer(y1 ~ time1 + (time1 | clus_id1),
           data = data.frame(y1, time1, clus_id1),
           REML = FALSE)
fs_dat1 <- get_fs_lmer(m1)
# Empirical Bayes
m2 <- lmer(y2 ~ time2 + (time2 | clus_id2),
           data = data.frame(y2, time2, clus_id2),
           REML = FALSE)
fs_dat2 <- get_fs_lmer(m2)
```

`get_fs_lmer()` carries each EB score (`u0_eb`, `u1_eb`) alongside its
per-cluster standard error (`u0_eb_se`), implied loadings (`u0_by_u0_eb`, …),
error variances (`ev_u0_eb`) and error covariance (`ecov_u0_eb_u1_eb`):

```{r}
knitr::kable(head(fs_dat1), digits = 3)
```

The loading, error-variance and error-covariance column names are exactly the
*definition variables* `tspa_mx_model()` reads in the model below.

One caveat: because `get_fs_lmer()` does not attach a per-cluster score
intercept (`fsb`), stage 2 fixes the score intercepts at zero. That does not
affect the latent *covariances* compared below (a covariance is separable from
the mean), but it pins the latent *mean* estimates (`u0 + u1 ~ 1`) to the shrunk
EB score means, so read those means with that caveat.

## Using EB Without Accounting for Error

```{r}
eb_m1 <- ranef(m1)[[1]]
eb_m2 <- ranef(m2)[[1]]
cor(cbind(eb_m1, eb_m2))
cov(cbind(eb_m1, eb_m2))
```

The numbers are quite different from the generating values.

## 2S-PA

### Process 1

```{r}
# Unstructured model (latent names distinct from the score columns)
cross_load <- matrix(c("u0_by_u0_eb", "u0_by_u1_eb",
                       "u1_by_u0_eb", "u1_by_u1_eb"), nrow = 2) |>
    `dimnames<-`(list(c("u0_eb", "u1_eb"), c("u0", "u1")))
err_cov <- matrix(c("ev_u0_eb", "ecov_u0_eb_u1_eb",
                    "ecov_u0_eb_u1_eb", "ev_u1_eb"), nrow = 2) |>
    `dimnames<-`(rep(list(c("u0_eb", "u1_eb")), 2))
tspa1_mx <- tspa_mx_model(model = "u0 ~~ u1\nu0 + u1 ~ 1",
                          data = fs_dat1, fsL = cross_load, fsT = err_cov)
# Run OpenMx
tspa1_mx_fit <- mxRun(tspa1_mx)
# Summarize the results
summary(tspa1_mx_fit)
```

The estimated variances and covariances for $u_0$ and $u_1$ are similar to the ones in the mixed model.

```{r}
print("2S-PA covariance")
lat_cov(tspa1_mx_fit, c("u0", "u1"))
print("lme4 covariance")
as.matrix(VarCorr(m1)$clus_id1)
```

Notes: The formulation works best with ML estimation for `lmer()`; with REML, one may need to consider the contrast matrix.

<!-- For REML,

$$
K' y = K' X \gamma + K' Z u + K' e,
$$
with K'X = 0. So

$$
K' y = K' Z u + K' e.
$$ -->

### Process 2

```{r}
cross_load <- matrix(c("u0_by_u0_eb", "u0_by_u1_eb",
                       "u1_by_u0_eb", "u1_by_u1_eb"), nrow = 2) |>
    `dimnames<-`(list(c("u0_eb", "u1_eb"), c("u0", "u1")))
err_cov <- matrix(c("ev_u0_eb", "ecov_u0_eb_u1_eb",
                    "ecov_u0_eb_u1_eb", "ev_u1_eb"), nrow = 2) |>
    `dimnames<-`(rep(list(c("u0_eb", "u1_eb")), 2))
tspa2_mx <- tspa_mx_model(model = "u0 ~~ u1\nu0 + u1 ~ 1",
                          data = fs_dat2, fsL = cross_load, fsT = err_cov)
# Run OpenMx
tspa2_mx_fit <- mxRun(tspa2_mx)
# Summarize the results
summary(tspa2_mx_fit)
```

```{r}
print("2S-PA covariance")
lat_cov(tspa2_mx_fit, c("u0", "u1"))
print("lme4 covariance")
as.matrix(VarCorr(m2)$clus_id2)
```

### Combining Processes 1 and 2

First, combine the two data sets with EB estimates. We'll add suffix to separate the variables.

```{r}
colnames(fs_dat1) <- paste0(colnames(fs_dat1), "_1")
colnames(fs_dat2) <- paste0(colnames(fs_dat2), "_2")
fs_dat <- cbind(fs_dat1, fs_dat2)
```

```{r}
# Unstructured model on the four growth-parameter latents
cross_load <- matrix(c("u0_by_u0_eb_1", "u0_by_u1_eb_1", NA, NA,
                       "u1_by_u0_eb_1", "u1_by_u1_eb_1", NA, NA,
                       NA, NA, "u0_by_u0_eb_2", "u0_by_u1_eb_2",
                       NA, NA, "u1_by_u0_eb_2", "u1_by_u1_eb_2"), nrow = 4) |>
    `dimnames<-`(list(c("u0_eb_1", "u1_eb_1", "u0_eb_2", "u1_eb_2"),
                      c("u0_1", "u1_1", "u0_2", "u1_2")))
err_cov <- matrix(c("ev_u0_eb_1", "ecov_u0_eb_u1_eb_1", NA, NA,
                    "ecov_u0_eb_u1_eb_1", "ev_u1_eb_1", NA, NA,
                    NA, NA, "ev_u0_eb_2", "ecov_u0_eb_u1_eb_2",
                    NA, NA, "ecov_u0_eb_u1_eb_2", "ev_u1_eb_2"), nrow = 4) |>
    `dimnames<-`(rep(list(c("u0_eb_1", "u1_eb_1", "u0_eb_2", "u1_eb_2")), 2))
tspa3_mx <- tspa_mx_model(
    model = "u0_1 ~~ u1_1 + u0_2 + u1_2\n
             u1_1 ~~ u0_2 + u1_2\n
             u0_2 ~~ u1_2\n
             u0_1 + u1_1 + u0_2 + u1_2 ~ 1",
    data = fs_dat, fsL = cross_load, fsT = err_cov)
# Run OpenMx
tspa3_mx_fit <- mxRun(tspa3_mx)
# Summarize the results
summary(tspa3_mx_fit)
```

```{r}
lat3 <- c("u0_1", "u1_1", "u0_2", "u1_2")
knitr::kable(cov(cbind(eb_m1, eb_m2)), digits = 3, caption = "EB covariance")
knitr::kable(lat_cov(tspa3_mx_fit, lat3), digits = 3, caption = "2S-PA covariance")
knitr::kable(latent_cov, digits = 3, caption = "Population covariance")
```

One can see the 2S-PA covariance matrix is much closer to the population values than using just EB estimates.

With only `n_clus = 100` clusters, any single draw carries substantial
sampling variability (a Monte-Carlo study over the same design gives standard
deviations of roughly `0.05`--`0.26` on these entries). The 2S-PA estimates are
essentially unbiased -- their mean across many replications recovers the
population values to within a few percent -- so the small gaps remaining here
are sampling error from one realization, not a bias in the estimator.

# With Missing Data

Now consider data with missing time points. We consider missing at random, where data are more likely to be missing for later time points.

```{r}
set.seed(1942)
# Missing indicators
r1 <- rbinom(time1, size = 1, prob = (20 - time1) / 20 - 0.05)
r2 <- rbinom(time2, size = 1, prob = (100 - time2) / 100 - 0.05)
```

Model with missing data

```{r}
# Empirical Bayes
m1 <- lmer(y1 ~ time1 + (time1 | clus_id1),
           data = data.frame(y1, time1, clus_id1),
           REML = FALSE, subset = r1 == 1)
fs_dat1 <- get_fs_lmer(m1)
# Empirical Bayes
m2 <- lmer(y2 ~ time2 + (time2 | clus_id2),
           data = data.frame(y2, time2, clus_id2),
           REML = FALSE, subset = r2 == 1)
fs_dat2 <- get_fs_lmer(m2)
colnames(fs_dat1) <- paste0(colnames(fs_dat1), "_1")
colnames(fs_dat2) <- paste0(colnames(fs_dat2), "_2")
fs_dat <- cbind(fs_dat1, fs_dat2)
```

Note that the cross-loadings and error covariances are not constant across clusters.

```{r}
head(fs_dat)
```

Now run 2S-PA and compare with EB estimates without accounting for error

```{r}
tspa3o_mx <- tspa_mx_model(
    model = "u0_1 ~~ u1_1 + u0_2 + u1_2\n
             u1_1 ~~ u0_2 + u1_2\n
             u0_2 ~~ u1_2\n
             u0_1 + u1_1 + u0_2 + u1_2 ~ 1",
    data = fs_dat, fsL = cross_load, fsT = err_cov)
# Run OpenMx
tspa3o_mx_fit <- mxTryHard(tspa3o_mx)
# Summarize the results
summary(tspa3o_mx_fit)
```

```{r}
lat3 <- c("u0_1", "u1_1", "u0_2", "u1_2")
knitr::kable(cov(cbind(fs_dat1[, 1:2], fs_dat2[, 1:2])), digits = 3, caption = "EB covariance")
knitr::kable(lat_cov(tspa3o_mx_fit, lat3), digits = 3, caption = "2S-PA covariance")
knitr::kable(latent_cov, digits = 3, caption = "Population covariance")
```
