---
title: "Multi-Factor Measurement Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Factor Measurement Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message = FALSE}
library(lavaan)
library(R2spa)
```

The example is from https://lavaan.ugent.be/tutorial/sem.html. 

## Factor score

For a CFA model with multiple latent factors, even when each indicator only loads on one factor, the resulting factor scores generally are weighted composites of **ALL** indicators. Consider the regression score, which has the form

$$\tilde{\boldsymbol \eta} = \mathbf{A}(\mathbf{y} - \hat{\boldsymbol \mu}) + \boldsymbol{\alpha}$$

where $\mathbf{A} = \boldsymbol{\Psi}\boldsymbol{\Lambda}^\top \hat{\boldsymbol{\Sigma}}^{-1}$ is a $q$ $\times$ $p$ matrix, $\hat{\boldsymbol \mu} = \boldsymbol{\nu} + \boldsymbol{\Lambda} \boldsymbol{\alpha}$ and $\boldsymbol{\Sigma} = \boldsymbol{\Lambda} \boldsymbol{\Psi}\boldsymbol{\Lambda}^\top + \boldsymbol{\Theta}$ are the model-implied means and covariances of the indicators $\mathbf{y}$, and $\boldsymbol{\alpha}$ and $\boldsymbol{\Psi}$ are the latent means and latent covariances. 

Therefore, assuming that the model is correctly specified such that $\mathbf{y} = \boldsymbol{\nu} + \boldsymbol{\Lambda} \boldsymbol{\eta} + \boldsymbol{\varepsilon}$,

$$
  \begin{aligned}
  \tilde{\boldsymbol \eta} & = \mathbf{A}(\boldsymbol{\Lambda} \boldsymbol{\eta} + \boldsymbol{\varepsilon} - \boldsymbol{\Lambda} \boldsymbol{\alpha}) + \boldsymbol{\alpha} \\
  & = (\mathbf{I} - \mathbf{A}\boldsymbol{\Lambda}) \boldsymbol{\alpha} +
  \mathbf{A}\boldsymbol{\Lambda} \boldsymbol{\eta} + \mathbf{A}\boldsymbol{\varepsilon}.
  \end{aligned}
$$

If we consider $\tilde{\boldsymbol \eta}$ as indicators of $\boldsymbol \eta$, we can see that $\boldsymbol{\nu}_\tilde{\boldsymbol{\eta}} = (\mathbf{I} - \mathbf{A}\boldsymbol{\Lambda}) \boldsymbol{\alpha}$ is the intercept, $\boldsymbol{\Lambda}_\tilde{\boldsymbol{\eta}} = \mathbf{A}\boldsymbol{\Lambda}$ is the $q$ $\times$ $q$ loading matrix, and $\boldsymbol{\Theta}_\tilde{\boldsymbol{\eta}} = \mathbf{A}\boldsymbol{\varepsilon}$ is the error covariance matrix.

We can see that $\boldsymbol{\Lambda}_\tilde{\boldsymbol{\eta}}$ is generally not diagonal, as the following shows:

```{r}
# CFA
my_cfa <- "
  # latent variables
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
"
cfa_fit <- cfa(model = my_cfa,
               data  = PoliticalDemocracy,
               std.lv = TRUE)
# A matrix
pars <- lavInspect(cfa_fit, what = "est")
lambda_mat <- pars$lambda
psi_mat <- pars$psi
sigma_mat <- cfa_fit@implied$cov[[1]]
ginvsigma <- MASS::ginv(sigma_mat)
alambda <- psi_mat %*% crossprod(lambda_mat, ginvsigma %*% lambda_mat)
alambda
```

We can also use `R2spa::get_fs()`:

```{r}
(fs_dat <- get_fs(PoliticalDemocracy, model = my_cfa, std.lv = TRUE)) |> head()
```

Therefore, cross-loadings must be accounted for in stage 2. Because `fs_dat`
comes from `get_fs()`, `tspa()` reads the implied loading matrix `fsL` (which
here carries the cross-loadings) and the score error covariance `fsT` from the
result's attributes, so the `fsT`/`fsL` arguments no longer need to be passed
(they can still be passed explicitly if preferred).

```{r}
tspa_fit <- tspa(model = "dem60 ~ ind60", data = fs_dat)
cat(attr(tspa_fit, "tspaModel"))
parameterestimates(tspa_fit)
```

The non-diagonal `fsL` (cross-loadings) is what makes the score-based model
recover the joint-model structure, because each score is a weighted composite of
all the indicators.

## Three-factor model example

```{r}
# CFA
cfa_3fac <-  "
  # latent variables
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
    dem65 =~ y5 + y6 + y7 + y8
"
cfa_3fac_fit <- cfa(model = cfa_3fac,
                    data  = PoliticalDemocracy,
                    std.lv = TRUE)
# A matrix
pars <- lavInspect(cfa_3fac_fit, what = "est")
lambda_mat <- pars$lambda
psi_mat <- pars$psi
sigma_mat <- cfa_3fac_fit@implied$cov[[1]]
ginvsigma <- MASS::ginv(sigma_mat)
alambda <- psi_mat %*% crossprod(lambda_mat, ginvsigma %*% lambda_mat)
alambda
```

We can also use `R2spa::get_fs()`:

```{r}
(fs_dat_3fac <- get_fs(PoliticalDemocracy, model = cfa_3fac, std.lv = TRUE)) |>
  head()
```

As before, `tspa()` reads the cross-loadings (`fsL`) and the score error
covariance (`fsT`) from the `get_fs()` result.

```{r}
tspa_fit_3fac <- tspa(model = "dem60 ~ ind60
                               dem65 ~ ind60 + dem60",
              data = fs_dat_3fac)
cat(attr(tspa_fit_3fac, "tspaModel"))
parameterestimates(tspa_fit_3fac)
```


Compare to SEM:

```{r}
sem_3fac <- sem("
  # latent variables
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
    dem65 =~ y5 + y6 + y7 + y8
  # structural model
    dem60 ~ ind60
    dem65 ~ ind60 + dem60
  ",
  data = PoliticalDemocracy
)
standardizedsolution(sem_3fac)
```

The `lavaan::sem()` fit above is the **gold-standard** joint model: it estimates
the measurement model and the structural paths together, treating the indicators
as observed, so no separate measurement-error correction is needed. `tspa()`
reaches the same structural estimates a different way — by correcting the stage-1
*scores* for their measurement error — so the two agree when the sample is large.
The simulation studies find 2S-PA tracks the joint model closely while staying
more reliable, with less biased standard errors and better coverage, in small
samples and with unreliable or categorical items (Lai & Hsiao, 2022; Lai et
al., 2023). `PoliticalDemocracy` is a real data set, so there is no "true value"
to compare against here; the point of putting the two side by side is that they
target the same structural model.

## Local vs joint scores

The `get_fs()` calls above all score from a *joint* multi-factor measurement
model: one fit with all the latents estimated together. With freely correlated
factors, each score is a weighted composite of **all** the indicators (the
off-diagonals of the `alambda` matrices above), so the stage-2 model has to
account for the cross-loadings explicitly. The canonical 2S-PA setup (Lai &
Hsiao, 2022) instead scores each latent from its **own** local measurement
model. `get_fs(..., local = TRUE)` does this in one call and merges the
per-construct results into the usual multi-factor layout:

```{r}
(fs_local <- get_fs(PoliticalDemocracy,
                    model = cfa_3fac,
                    std.lv = TRUE,
                    local = TRUE)) |> head()
```

The merged result has the same layout as the joint output, with the cross
terms *exactly* zero: the `fsT` and `fsL` attributes are block-diagonal, the
`psi` attribute is diagonal, and the off-diagonal `_by_` loading columns and
all `ecov_*` columns are zero. Each score is a *pure* per-construct score —
the cross-factor structure is not estimated by design:

```{r}
attr(fs_local, "fsL")[[1]]
sapply(c("fs_ind60", "fs_dem60", "fs_dem65"),
       function(col) max(abs(fs_local[[col]] - fs_dat_3fac[[col]])))
```

So local scores are not the same as joint scores: with freely correlated
factors they differ (the maximum score differences above), because the joint
fit scores every latent from all the indicators — even the joint fit's own
per-factor estimates shift under free correlation. With the factors
constrained uncorrelated (e.g. `ind60 ~~ 0 * dem60` in the joint model), the
likelihood factorizes and the two agree to optimizer tolerance.

Because the merged result carries the usual `fsT`/`fsL` attributes (with the
zero cross terms), it feeds `tspa()` directly with no explicit `fsT`/`fsL`
arguments — the stage-2 model picks up the zero cross-loadings and score
error covariances from the attributes:

```{r}
tspa_fit_local <- tspa(model = "dem60 ~ ind60
                        dem65 ~ ind60 + dem60",
                       data = fs_local)
parameterestimates(tspa_fit_local)
```

`model` may also be a character vector (or a named list) of complete
single-factor model strings, each element fit verbatim — the escape hatch for
anything the strict string grammar rejects, e.g. a within-factor residual
covariance (`y1 ~~ y4`). `vfsLT` (and hence `tspa(corrected_se = TRUE)`),
`prior_cov`, and `reliability` are not supported in `local` mode (v1); see
`?get_fs` (Details) for the full contract.
