---
title: "Multi-Factor Measurement Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multiple-factors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message = FALSE}
library(lavaan)
library(R2spa)
```

The example is from https://lavaan.ugent.be/tutorial/sem.html. 

## Factor score

For a CFA model with multiple latent factors, even when each indicator only loads on one factor, the resulting factor scores generally are weighted composites of **ALL** indicators. Consider the regression score, which has the form

$$\tilde{\boldsymbol \eta} = \mathbf{A}(\mathbf{y} - \hat{\boldsymbol \mu}) + \boldsymbol{\alpha}$$

where $\mathbf{A} = \boldsymbol{\Psi}\boldsymbol{\Lambda}^\top \hat{\boldsymbol{\Sigma}}^{-1}$ is a $q$ $\times$ $p$ matrix, $\hat{\boldsymbol \mu} = \boldsymbol{\nu} + \boldsymbol{\Lambda} \boldsymbol{\alpha}$ and $\boldsymbol{\Sigma} = \boldsymbol{\Lambda} \boldsymbol{\Psi}\boldsymbol{\Lambda}^\top + \boldsymbol{\Theta}$ are the model-implied means and covariances of the indicators $\mathbf{y}$, and $\boldsymbol{\alpha}$ and $\boldsymbol{\Psi}$ are the latent means and latent covariances. 

Therefore, assuming that the model is correctly specified such that $\mathbf{y} = \boldsymbol{\nu} + \boldsymbol{\Lambda} \boldsymbol{\eta} + \boldsymbol{\varepsilon}$,

$$
  \begin{aligned}
  \tilde{\boldsymbol \eta} & = \mathbf{A}(\boldsymbol{\Lambda} \boldsymbol{\eta} + \boldsymbol{\varepsilon} - \boldsymbol{\Lambda} \boldsymbol{\alpha}) + \boldsymbol{\alpha} \\
  & = (\mathbf{I} - \mathbf{A}\boldsymbol{\Lambda}) \boldsymbol{\alpha} +
  \mathbf{A}\boldsymbol{\Lambda} \boldsymbol{\eta} + \mathbf{A}\boldsymbol{\varepsilon}.
  \end{aligned}
$$

If we consider $\tilde{\boldsymbol \eta}$ as indicators of $\boldsymbol \eta$, we can see that $\boldsymbol{\nu}_\tilde{\boldsymbol{\eta}} = (\mathbf{I} - \mathbf{A}\boldsymbol{\Lambda}) \boldsymbol{\alpha}$ is the intercept, $\boldsymbol{\Lambda}_\tilde{\boldsymbol{\eta}} = \mathbf{A}\boldsymbol{\Lambda}$ is the $q$ $\times$ $q$ loading matrix, and $\boldsymbol{\Theta}_\tilde{\boldsymbol{\eta}} = \mathbf{A}\boldsymbol{\varepsilon}$ is the error covariance matrix.

We can see that $\boldsymbol{\Lambda}_\tilde{\boldsymbol{\eta}}$ is generally not diagonal, as the following shows:

```{r}
# CFA
my_cfa <- "
  # latent variables
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
"
cfa_fit <- cfa(model = my_cfa,
               data  = PoliticalDemocracy,
               std.lv = TRUE)
# A matrix
pars <- lavInspect(cfa_fit, what = "est")
lambda_mat <- pars$lambda
psi_mat <- pars$psi
sigma_mat <- cfa_fit@implied$cov[[1]]
ginvsigma <- MASS::ginv(sigma_mat)
alambda <- psi_mat %*% crossprod(lambda_mat, ginvsigma %*% lambda_mat)
alambda
```

We can also use `R2spa::get_fs()`:

```{r}
(fs_dat <- get_fs(PoliticalDemocracy, model = my_cfa, std.lv = TRUE)) |> head()
```

Therefore, we need to specify cross-loadings when using 2S-PA.

```{r}
tspa_fit <- tspa(model = "dem60 ~ ind60", data = fs_dat,
                 fsT = attr(fs_dat, "fsT"), 
                 fsL = attr(fs_dat, "fsL"))
cat(attr(tspa_fit, "tspaModel"))
parameterestimates(tspa_fit)
```


```{r include=FALSE, eval=FALSE}
# Compute factor scores using lavPredict
fs_dat <- lavPredict(cfa_fit, acov = "standard")
colnames(fs_dat) <- c("fs_ind60", "fs_dem60")
# alambda %*% attr(fs_dat, "acov")[[1]] %*% t(alambda)
my2spa <- "
  # latent variables (indicated by factor scores)
    ind60 =~ 0.95538579 * fs_ind60 + 0.05816994 * fs_dem60
    dem60 =~ 0.01834111 * fs_ind60 + 0.86888887 * fs_dem60
  # constrain the errors
    fs_ind60 ~~ 0.034595518 * fs_ind60
    fs_dem60 ~~ 0.090775708 * fs_dem60
    fs_ind60 ~~ 0.004017388 * fs_dem60
  # latent variances
    ind60 ~~ ind60
    dem60 ~~ dem60
  # regressions
    dem60 ~ ind60
"
tspa_fit <- sem(model = my2spa, data  = fs_dat)
standardizedsolution(tspa_fit)
```

which is more consistent with the SEM results.

## Three-factor model example

```{r}
# CFA
cfa_3fac <-  "
  # latent variables
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
    dem65 =~ y5 + y6 + y7 + y8
"
cfa_3fac_fit <- cfa(model = cfa_3fac,
                    data  = PoliticalDemocracy,
                    std.lv = TRUE)
# A matrix
pars <- lavInspect(cfa_3fac_fit, what = "est")
lambda_mat <- pars$lambda
psi_mat <- pars$psi
sigma_mat <- cfa_3fac_fit@implied$cov[[1]]
ginvsigma <- MASS::ginv(sigma_mat)
alambda <- psi_mat %*% crossprod(lambda_mat, ginvsigma %*% lambda_mat)
alambda
```

We can also use `R2spa::get_fs()`:

```{r}
(fs_dat_3fac <- get_fs(PoliticalDemocracy, model = cfa_3fac, std.lv = TRUE)) |>
  head()
```

Therefore, we need to specify cross-loadings when using 2S-PA.

```{r}
tspa_fit_3fac <- tspa(model = "dem60 ~ ind60
              dem65 ~ ind60 + dem60",
              data = fs_dat_3fac,
              fsT = attr(fs_dat_3fac, "fsT"),
              fsL = attr(fs_dat_3fac, "fsL"))
cat(attr(tspa_fit_3fac, "tspaModel"))
parameterestimates(tspa_fit_3fac)
```


```{r include=FALSE, eval=FALSE}
# Compute factor scores using lavPredict
fs_3fac_dat <- lavPredict(cfa_3fac_fit, acov = "standard")
colnames(fs_3fac_dat) <- c("fs_ind60", "fs_dem60", "fs_dem65")
# alambda %*% attr(fs_dat, "acov")[[1]] %*% t(alambda)
mod_2spa_3fac <- "
  # latent variables (indicated by factor scores)
    ind60 =~ 0.95064774 * fs_ind60 + (-0.02069724) * fs_dem60 + 
             0.09868528 * fs_dem65
    dem60 =~ (-0.005967124) * fs_ind60 + 0.533020047 * fs_dem60 +
             0.401095944 * fs_dem65
    dem65 =~ 0.02951139 * fs_ind60 + 0.41603787 * fs_dem60 +
             0.49133377 * fs_dem65
  # constrain the errors
    fs_ind60 ~~ 0.0340104952 * fs_ind60 + 0.0003881641 * fs_dem60 + 
                0.005026024 * fs_dem65
    fs_dem60 ~~ 0.0586870314 * fs_dem60 + 0.0533752458 * fs_dem65
    fs_dem65 ~~ 0.051573299 * fs_dem65
  # latent variances
    ind60 ~~ ind60
    dem60 ~~ dem60
    dem65 ~~ dem65
  # regressions
    dem60 ~ ind60
    dem65 ~ ind60 + dem60
"
tspa_fit_3fac <- sem(model = mod_2spa_3fac, data = fs_3fac_dat)
standardizedsolution(tspa_fit_3fac)
```

Compare to SEM:

```{r}
sem_3fac <- sem("
  # latent variables
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
    dem65 =~ y5 + y6 + y7 + y8
  # structural model
    dem60 ~ ind60
    dem65 ~ ind60 + dem60
  ",
  data = PoliticalDemocracy
)
standardizedsolution(sem_3fac)
```
