---
title: "Two-Stage Path Analysis (2S-PA) Model Examples"
author: "Yixiao Li"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{2spa_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The example is from https://lavaan.ugent.be/tutorial/sem.html. 

Load packages

```{r load-pkg, message = FALSE}
library(lavaan)
library(R2spa)
```

## Why two-stage path analysis

When the constructs in a path model are measured with error, plugging composite
or factor scores into a regression as if they were error-free biases the path
coefficients (Cole & Preacher, 2014). A single joint structural equation model,
estimating the
measurement and structural models together, is the gold standard, but it can be
hard to specify, is prone to convergence and admissibility problems in small
samples, and forces a whole-model refit whenever the measurement *or* the
structural part changes (Rosseel, 2012).

Two-stage path analysis (2S-PA; Lai & Hsiao, 2022) is a practical alternative
that splits the problem:

- **Stage 1 — score each construct** with an appropriate psychometric model (a
  `lavaan` CFA, an `lme4` mixed model, a `mirt` IRT model, or even an EFA) and
  record each score's measurement error / reliability.
- **Stage 2 — regress on the scores**, correcting for each score's measurement
  error with *definition variables* that fix the error variance to its
  observation-specific value.

Because the stages are separate, you can tune and diagnose each on its own, mix
estimators across constructs (one construct from a CFA, another from an IRT
model), and avoid the small-sample convergence problems of one large joint
model. Simulations find 2S-PA tracks the joint model in large samples but
converges more often, with less standard-error bias and better Type-I error /
CI coverage in small samples and with categorical items (Lai & Hsiao, 2022;
Lai et al., 2023); the same two-stage idea also covers empirical-Bayes random
slopes from mixed models (Lai & Liu, 2026).

`R2spa` implements both stages: `get_fs()` is stage 1 (scores plus their
loadings, error covariances and intercepts) and `tspa()` / `tspa_mx_model()`
are stage 2.

## How the measurement error is supplied

`tspa()` needs each score's measurement error, in one of two forms:

- **`se_fs`** — one standard error per score (a scalar, or a list / data frame
  of per-group scalars for multigroup). This treats the error *variance* as
  constant across observations and assumes each score is an indicator of a
  single latent (no cross-loadings, no correlated score errors). It is the
  simple path: suitable for a unidimensional construct scored from continuous,
  roughly equally reliable items.
- **`fsT`** (with **`fsL`** and **`fsb`**) — the general form `get_fs()` returns
  as attributes (`attr(fs, "fsT")`, `attr(fs, "fsL")`, `attr(fs, "fsb")`).
  `fsT` is the score error *variance-covariance* (so it also carries correlated
  errors and cross-loadings) and `fsL` the implied loading of each latent on the
  score. Use this when scores are composite (several indicators per latent, or a
  score loading on several latents), and note it is the only form that carries
  per-observation reliability, which the `mirt` (IRT) and `merMod` (mixed
  model) paths need because reliability varies by person / cluster (see the
  `multiple-factors`, `tspa-vignette-mx` and `multilevel` vignettes).

The examples below use `se_fs`, since each construct is a single-factor score
from continuous items; the `multiple-factors` vignette shows the `fsT`/`fsL`
form.

## Single group, single predictor^[In the model, `a`, `b`, and `c` are coefficients where variables `y1`, `y2`, `y3`, and `y4` are indicators of the latent variable `dem60`.]

```{r}
model <- ' 
  # latent variable definitions
     ind60 =~ x1 + x2 + x3
     dem60 =~ y1 + y2 + y3 + y4

  # regressions
    dem60 ~ ind60
'
```

To call `tspa()`, a data frame of factor scores is needed for all latent variables. To get this data frame, apply `get_fs()` to all latent variables and specify model parameter as their respective definitions. Combine factor scores for all latent variable using `cbind()` so that it can be used in `tspa()` for model building.

```{r}
fs_dat_ind60 <- get_fs(PoliticalDemocracy, 
                       model = "ind60 =~ x1 + x2 + x3")
fs_dat_dem60 <- get_fs(PoliticalDemocracy, 
                       model = "dem60 =~ y1 + y2 + y3 + y4")
fs_dat <- cbind(fs_dat_ind60, fs_dat_dem60)
```

To build a Two-Stage Path Analysis model, simply call the `tspa()` function with model = regressions, data = The combined factor score data frame using `get_fs()`, and specify standard error as either a list or a data frame. Values for standard error can be found at column named `fs_[variable name]_se`. For example, the standard error of latent variable ind60 can be found at column `fs_ind60_se` in the `fs_dat` data frame.

```{r}
tspa_fit <- tspa(model = "dem60 ~ ind60", 
                 data = fs_dat, 
                 se_fs = list(ind60 = 0.1213615, dem60 = 0.6756472))
```

To view the Two-Stage Path Analysis model, use `attributes([model name])$tspaModel`. Function `cat()` can help tidy up the model output. In the output, the values of error constraints are computed by squaring standard errors from the previous section.

```{r}
cat(attributes(tspa_fit)$tspaModel)
```

## Single group, multiple predictors

```{r}
model <- ' 
  # latent variable definitions
     ind60 =~ x1 + x2 + x3
     dem60 =~ y1 + y2 + y3 + y4
     dem65 =~ y5 + y6 + y7 + y8

  # regressions
    dem60 ~ ind60
    dem65 ~ ind60 + dem60

  # # residual correlations
  #   y1 ~~ y5
  #   y2 ~~ y4 + y6
  #   y3 ~~ y7
  #   y4 ~~ y8
  #   y6 ~~ y8
'
```

To call `tspa()`, a data frame of factor scores is needed for all latent variables. To get this data frame, apply `get_fs()` to all latent variables and specify model parameters as their respective definitions. Combine factor scores for all latent variables using `cbind()` to call `tspa()` for model building.

```{r}
fs_dat_ind60 <- get_fs(PoliticalDemocracy, 
                       model = "ind60 =~ x1 + x2 + x3")
fs_dat_dem60 <- get_fs(PoliticalDemocracy, 
                       model = "dem60 =~ y1 + y2 + y3 + y4")
fs_dat_dem65 <- get_fs(PoliticalDemocracy, 
                       model = "dem65 =~ y5 + y6 + y7 + y8")
fs_dat <- cbind(fs_dat_ind60, fs_dat_dem60, fs_dat_dem65)
```

To build a Two-Stage Path Analysis model, simply call the `tspa()` function with model = regressions (for all predictors), data = the factor score data frame created by combining results from `get_fs()`, and specify standard errors as either a list or a data frame. Values for standard errors can be found at column named `fs_[variable name]_se`.^[The text inside `[]` should be replaced with actual variable and model name.] For example, the standard error of latent variable ind60 can be found at column `fs_ind60_se` in the `fs_dat` data frame. In the output model, the values of error constraints are computed by squaring standard errors.

```{r}
tspa_3var_fit <- tspa(model = "dem60 ~ ind60
                          dem65 ~ ind60 + dem60", 
                      data = fs_dat, 
                      se_fs = list(ind60 = 0.1213615, dem60 = 0.6756472, 
                                   dem65 = 0.5724405))
cat(attributes(tspa_3var_fit)$tspaModel)
```

## Multigroup, single predictor

```{r}
model <- ' 
  # latent variable definitions
    visual =~ x1 + x2 + x3
    speed =~ x7 + x8 + x9

  # regressions
    visual ~ speed
'
```

To call `tspa()`, a data frame of factor scores is needed for all multigroup variables. To get this data frame, apply `get_fs()` to all multigroup variables and specify model parameter as their respective definitions. Combine factor scores for all multigroup variables using `cbind()` so that it can be fed in `tspa()` for model building.

```{r}
hs_mod <- '
visual =~ x1 + x2 + x3
speed =~ x7 + x8 + x9
'

# get factor scores
fs_dat_visual <- get_fs(HolzingerSwineford1939, 
                        model = "visual =~ x1 + x2 + x3", 
                        group = "school")
fs_dat_speed <- get_fs(HolzingerSwineford1939, 
                       model = "speed =~ x7 + x8 + x9", 
                       group = "school")
fs_hs <- cbind(fs_dat_visual, fs_dat_speed)
```

To build a Two-Stage Path Analysis model, simply call the `tspa()` function with model = regression relation. Specify standard error as either a list or a data frame. Values for standard error can be found at column named `fs_[variable name]_se`. For example, the standard error of the multigroup variable visual can be found at column `fs_visual_se` in the `fs_hs` data frame. To get the standard errors for each group faster, `unique()` can be called upon the standard error column. For example, in this case, `unique(fs_hs$fs_visual_se)` can be called to get the standard errors for multigroup variable visual.

Function `standardizedsolution()` enables user to view a table of standard error, z score, p-value, and lower bound of confidence interval for each multigroup regression relation.

```{r}
# tspa model
tspa_fit <- tspa(model = "visual ~ speed",
                 data = fs_hs,
                 se_fs = data.frame(visual = c(0.3391326, 0.311828),
                                    speed = c(0.2786875, 0.2740507)),
                 group = "school"
                 # group.equal = "regressions"
                 )
stdsol <- standardizedsolution(tspa_fit)
subset(stdsol, subset = op == "~")
```

To view the Two-Stage Path Analysis model for multigroup, use `attributes([model name])$tspaModel`. Function `cat()` can help tidy up the model output. In the output, the values of error constraints are computed by squaring standard errors from the previous section.

```{r}
cat(attributes(tspa_fit)$tspaModel)
```

## References

- Cole, D. A., & Preacher, K. J. (2014). Manifest variable path analysis:
  Potentially serious and misleading consequences due to uncorrected measurement
  error. *Psychological Methods, 19*(2), 300–315.
  [https://doi.org/10.1037/a0033805](https://doi.org/10.1037/a0033805)
- Lai, M. H. C., & Hsiao, Y.-Y. (2022). Two-stage path analysis with definition
  variables: An alternative framework to account for measurement error.
  *Psychological Methods, 27*(4), 568–588.
  [https://doi.org/10.1037/met0000410](https://doi.org/10.1037/met0000410)
- Lai, M. H. C., Tse, W. W.-Y., Zhang, G., Li, Y., & Hsiao, Y.-Y. (2023).
  Correcting for unreliability and partial invariance: A two-stage path analysis
  approach. *Structural Equation Modeling: A Multidisciplinary Journal, 30*(2),
  258–271.
  [https://doi.org/10.1080/10705511.2022.2125397](https://doi.org/10.1080/10705511.2022.2125397)
- Lai, M. H. C., & Liu, S. (2026). A two-stage approach to account for
  measurement error when using empirical Bayes estimates of random slopes.
  *Psychological Methods.* Advance online publication.
  [https://doi.org/10.1037/met0000838](https://doi.org/10.1037/met0000838)
- Rosseel, Y. (2012). *lavaan*: An R package for structural equation modeling.
  *Journal of Statistical Software, 48*(2), 1–36.
  [https://doi.org/10.18637/jss.v048.i02](https://doi.org/10.18637/jss.v048.i02)
