---
title: "Two-Stage Path Analysis (2S-PA) Model Examples"
author: "Yixiao Li"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Two-Stage Path Analysis (2S-PA) Model Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The example is from https://lavaan.ugent.be/tutorial/sem.html. 

Load packages

```{r load-pkg, message = FALSE}
library(lavaan)
library(R2spa)
```

## Why two-stage path analysis

When the constructs in a path model are measured with error, plugging composite
or factor scores into a regression as if they were error-free biases the path
coefficients (Cole & Preacher, 2014). A single joint structural equation model,
estimating the
measurement and structural models together, is the gold standard, but it can be
hard to specify, is prone to convergence and admissibility problems in small
samples, and forces a whole-model refit whenever the measurement *or* the
structural part changes (Rosseel, 2012).

Two-stage path analysis (2S-PA; Lai & Hsiao, 2022) is a practical alternative
that splits the problem:

- **Stage 1 — score each construct** with an appropriate psychometric model (a
  `lavaan` CFA, an `lme4` mixed model, a `mirt` IRT model, or even an EFA) and
  record each score's measurement error / reliability.
- **Stage 2 — path-analyze the scores**, correcting for measurement
  error by fixing the loading(s) and error variance (and covariances) to values
  implied in Stage 1. Observation-specific values can be used via 
  *definition variables* (currently available in *OpenMx*).

Because the stages are separate, you can tune and diagnose each on its own, mix
estimators across constructs (one construct from a CFA, another from an IRT
model), and avoid the small-sample convergence problems of one large joint
model. Simulations find 2S-PA tracks the joint model in large samples but
converges more often, with less standard-error bias and better Type-I error /
CI coverage in small samples and with categorical items (Lai & Hsiao, 2022;
Lai et al., 2023); the same two-stage idea also covers empirical-Bayes random
slopes from mixed models (Lai & Liu, 2026).

`R2spa` implements both stages: `get_fs()` is stage 1 (scores plus their
loadings, error covariances and intercepts) and `tspa()` / `tspa_mx_model()`
are stage 2.

## Why the standardized coefficients are of interest

The latent constructs in a 2S-PA — like the latents in any latent-variable
model — have **no intrinsic scale**. A latent variable is not directly
observed; its units are arbitrary and are fixed only by an identification
constraint in the measurement model (a loading pinned to 1, or a variance
pinned to 1 via `std.lv = TRUE`). The unstandardized stage-2 path coefficients
therefore inherit that arbitrary scale (units of latent `dem60` per unit of
latent `ind60`), and their magnitudes shift if the latent is rescaled.

The **standardized** coefficients are dimensionless — each path is the effect
of a one-SD change in the predictor on the outcome, in SD units — and are
invariant to the arbitrary choice of latent metric. They are therefore usually
the parameters of substantive interest, and they are comparable across
constructs and between 2S-PA and the joint SEM (the two agree when the sample
is large). Inspect them with `summary(fit, standardized = TRUE)` or
`lavaan::standardizedSolution(fit)`. For multigroup fits, `standardizedSolution()`
standardizes within group and `grandStandardizedSolution()` pools across groups
(see the [Grand Standardized Coefficients](gr-std-coef.html) vignette).

## How the measurement error is supplied

`tspa()` needs each score's measurement error, in one of two forms:

- **`se_fs`** — one standard error per score (a scalar, or a list / data frame
  of per-group scalars for multigroup). This treats the error *variance* as
  constant across observations and assumes each score is an indicator of a
  single latent (no cross-loadings, no correlated score errors). It is the
  simple path: suitable for a unidimensional construct scored from continuous,
  roughly equally reliable items.
- **`fsT`** (with **`fsL`** and **`fsb`**) — the general form `get_fs()` returns
  as attributes (`attr(fs, "fsT")`, `attr(fs, "fsL")`, `attr(fs, "fsb")`).
  `fsT` is the score error *variance-covariance* (so it also carries correlated
  errors and cross-loadings) and `fsL` the implied loading of each latent on the
  score. Use this when scores are composite (several indicators per latent, or a
  score loading on several latents), and note it is the only form that carries
  per-observation measurement properties, which the `mirt` (IRT) and `merMod` (mixed
  model) paths need because loadings and error variances vary by person / cluster (see the
  `multiple-factors`, `tspa-vignette-mx` and `multilevel` vignettes).

Each construct below is a single-factor score from continuous items, so
`se_fs` is the natural form: the single-factor example passes it explicitly,
while the multi-factor and multigroup examples omit it, since `tspa()` reads
those standard errors from the `get_fs()` result automatically. The
`multiple-factors` vignette shows the `fsT`/`fsL` form.
The `OpenMx` backend `tspa_mx_model()` likewise omits the
measurement inputs on a `get_fs()` result, deriving them from the
result's attributes (per-row quantities as definition variables —
the exact, non-pooled route; see the `tspa-vignette-mx` vignette).

## Reading the `get_fs()` output

`get_fs()` returns one row per observation. For a construct called `<name>` it
contributes these columns, with the matching matrix-form quantities attached
as attributes on the returned data frame:

| Column(s) / attribute | Meaning |
| --- | --- |
| `fs_<name>` | the factor score |
| `fs_<name>_se` | that score's standard error — the measurement-error term `tspa()` corrects for |
| `ev_fs_<name>` | the score's error variance |
| `ecov_fs_<a>_fs_<b>` | error covariance between scores `a` and `b` (only from a shared multi-factor model) |
| `<indicator>_by_fs_<name>` | model-implied loading — the mean of the indicator on the score |
| `attr(fs, "fsT")` | the score error variance-covariance (carries correlated errors and cross-loadings) |
| `attr(fs, "fsL")` | the implied loading of each latent on the score |
| `attr(fs, "fsb")` | the scoring-equation intercept, one per score (named by the score, e.g. `fs_ind60`; `0` for a complete-data single-factor CFA, varying per row when the measurement quantities do) |

The attributes are what `tspa_mx_model()` consumes, and `tspa(fsb = ...)` uses
`fsb` to fix the stage-2 intercept constraints. `fs_indiv(include_intercept =
TRUE)` materializes the `fsb` values as `int_fs_<f>` columns, which is how the
`tspa_mx_model()` examples wire up the score-intercept definition variables
(`fsb = c(fs_ind60 = "int_fs_ind60", ...)`).

## Mean (sum) scores

`get_fs()` can also score a construct as its **mean score** — the plain mean of
the items that load on it, with no latent distribution — a simpler alternative
to the default regression (EB) scores. Request it with `method = "mean"` (the
item-to-factor split is auto-derived from the estimated loadings, or supplied
via `sum_items`):

```{r}
fs_mean <- get_fs(PoliticalDemocracy,
                  model = "ind60 =~ x1 + x2 + x3
                           dem60 =~ y1 + y2 + y3 + y4",
                  method = "mean")
head(fs_mean)   # fs_ind60 == rowMeans(x1, x2, x3); constant fs_<name>_se per factor
```

A mean score is a composite of items, so its measurement error is carried by
the `fsT`/`fsL`/`fsb` form (the cross terms are exactly zero for a plain CFA).
`tspa()` reads those from the result's attributes:

```{r}
tspa_mean_fit <- tspa(model = "dem60 ~ ind60",
                      data = fs_mean,
                      fsT = attr(fs_mean, "fsT"),
                      fsL = attr(fs_mean, "fsL"),
                      fsb = attr(fs_mean, "fsb"))
summary(tspa_mean_fit, standardized = TRUE)
```

`method = "mean"` is for complete-data fits: it errors when stage 1 retains
missing data (e.g. `missing = "fiml"`) and is not supported together with
`corrected_fsT`, `vfsLT`, `reliability`, `prior_mean`, or `prior_cov`.

## Single group, single factor

```r
model <- ' 
  # latent variable definitions
     ind60 =~ x1 + x2 + x3
     dem60 =~ y1 + y2 + y3 + y4

  # regressions
    dem60 ~ ind60
'
```

To call `tspa()`, a data frame of factor scores is needed for all latent variables. To get this data frame, apply `get_fs()` to all latent variables and specify model parameters as their respective definitions. Combine factor scores for all latent variables using `cbind()` so that it can be used in `tspa()` for model building.

```{r}
fs_dat_ind60 <- get_fs(PoliticalDemocracy, 
                       model = "ind60 =~ x1 + x2 + x3")
fs_dat_dem60 <- get_fs(PoliticalDemocracy, 
                       model = "dem60 =~ y1 + y2 + y3 + y4")
fs_dat <- cbind(fs_dat_ind60, fs_dat_dem60)
head(fs_dat)
```

To build a Two-Stage Path Analysis model, simply call the `tspa()` function with model = regressions, data = the combined factor score data frame using `get_fs()`, and specify standard error as either a list or a data frame. Values for standard error can be found at column named `fs_[variable name]_se`. For example, the standard error of latent variable ind60 can be found at column `fs_ind60_se` in the `fs_dat` data frame.

```r
tspa_fit <- tspa(model = "dem60 ~ ind60", 
                 data = fs_dat, 
                 se_fs = list(ind60 = 0.1213615, dem60 = 0.6756472))
```

We recently added support for `tspa()` to directly grep the `fs_[variable name]_se` columns from the `get_fs()` output, so the `se_fs` argument can be omitted:

```{r}
tspa_fit <- tspa(model = "dem60 ~ ind60", 
                 data = fs_dat)
summary(tspa_fit, standardized = TRUE)
```

The **standardized** coefficients (the `est.std` column) are usually the
parameters of interest, because the latent constructs have no intrinsic scale
(see the 'Why the standardized coefficients are of interest' section above).
To view the Two-Stage Path Analysis model, use `attributes([model name])$tspaModel`. Function `cat()` can help tidy up the model output. In the output, the values of error constraints are computed by squaring standard errors from the previous section.

```{r}
cat(attributes(tspa_fit)$tspaModel)
```

## Single group, multiple factors

```r
model <- ' 
  # latent variable definitions
     ind60 =~ x1 + x2 + x3
     dem60 =~ y1 + y2 + y3 + y4
     dem65 =~ y5 + y6 + y7 + y8

  # regressions
    dem60 ~ ind60
    dem65 ~ ind60 + dem60

  # # residual correlations
  #   y1 ~~ y5
  #   y2 ~~ y4 + y6
  #   y3 ~~ y7
  #   y4 ~~ y8
  #   y6 ~~ y8
'
```

To call `tspa()`, a data frame of factor scores is needed for all latent variables. To get this data frame, apply `get_fs()` to all latent variables and specify model parameters as their respective definitions. Combine factor scores for all latent variables using `cbind()` to call `tspa()` for model building.

```r
fs_dat_ind60 <- get_fs(PoliticalDemocracy, 
                       model = "ind60 =~ x1 + x2 + x3")
fs_dat_dem60 <- get_fs(PoliticalDemocracy, 
                       model = "dem60 =~ y1 + y2 + y3 + y4")
fs_dat_dem65 <- get_fs(PoliticalDemocracy, 
                       model = "dem65 =~ y5 + y6 + y7 + y8")
fs_dat <- cbind(fs_dat_ind60, fs_dat_dem60, fs_dat_dem65)
```

We recently made `get_fs()` to support a multi-factor model input, but computing the factor scores separately for each latent variable via the argument `local = TRUE`.

```{r}
# This is equivalent to the previous chunk:
fs_dat <- get_fs(PoliticalDemocracy, 
                 model = "ind60 =~ x1 + x2 + x3
                          dem60 =~ y1 + y2 + y3 + y4
                          dem65 =~ y5 + y6 + y7 + y8",
                 local = TRUE)
head(fs_dat)  # check that the error covariances and cross-loadings are zero.
```

To build a Two-Stage Path Analysis model, simply call the `tspa()` function with model = regressions (for all predictors) and data = the factor score data frame created by combining results from `get_fs()`. Each score's standard error is available in the column named `fs_[variable name]_se`.^[The text inside `[]` should be replaced with the actual variable name.] For example, the standard error of latent variable ind60 is in column `fs_ind60_se` of the `fs_dat` data frame. Because the combined frame carries each score's `fs_[variable name]_se` column from `get_fs()`, `tspa()` reads those standard errors from the columns automatically, so the `se_fs` argument can be omitted (passing it explicitly as a list or a data frame still works). In the output model, the values of error constraints are computed by squaring the standard errors.

```{r}
tspa_3var_fit <- tspa(model = "dem60 ~ ind60
                               dem65 ~ ind60 + dem60", 
                      data = fs_dat)
cat(attributes(tspa_3var_fit)$tspaModel)
```

## Multigroup, multiple factors

```r
model <- ' 
  # latent variable definitions
    visual =~ x1 + x2 + x3
    speed =~ x7 + x8 + x9

  # regressions
    visual ~ speed
'
```

To call `tspa()`, a data frame of factor scores is needed for all latent variables for each group. To get this data frame, you can use the above strategy to compute factor scores separately for each latent variable with the `local = TRUE` argument. Alternatively, you can use the default (`local = FALSE`), which computes factor scores for all latent variables simultaneously, but resulting in correlated measurement errors and cross-loadings with the regression method for factor scores (see more discussion in the [`Scoring Matrices: lavaan CFA and lme4`](scoring-matrices.html) vignette).

```{r}
hs_mod <- '
visual =~ x1 + x2 + x3
speed =~ x7 + x8 + x9
'

# get factor scores (joint model)
fs_hs <- get_fs(HolzingerSwineford1939, model = hs_mod, group = "school")
head(fs_hs)  # with cross-loadings and correlated measurement errors
```

To build a Two-Stage Path Analysis model, simply call the `tspa()` function with model = the regression relation, data = the combined factor score data frame, and `group` = the grouping column (here `"school"`). Each score's standard error is available in the column named `fs_[variable name]_se`. For example, the standard error of the multigroup variable visual is in column `fs_visual_se` of the `fs_hs` data frame. Because `fs_hs` comes from `get_fs()`, `tspa()` reads the per-group standard errors from those columns automatically, so the `se_fs` argument can be omitted (passing it explicitly as a data frame still works). The `unique()` function is handy for inspecting the per-group standard errors: for example, `unique(fs_hs$fs_visual_se)` shows the standard error of visual in each group.

Function `standardizedsolution()` enables user to view a table of the standardized estimate, standard error, z score, p-value, and confidence interval bounds for each multigroup regression relation.

```{r}
# tspa model
tspa_fit <- tspa(model = "visual ~ speed",
                 data = fs_hs,
                 group = "school"
                 # group.equal = "regressions"
                 )
stdsol <- standardizedsolution(tspa_fit)
subset(stdsol, subset = op == "~")
```

To view the Two-Stage Path Analysis model for multigroup, use `attributes([model name])$tspaModel`. Function `cat()` can help tidy up the model output. In the output, the values of error constraints are computed by squaring standard errors from the previous section.

```{r}
cat(attributes(tspa_fit)$tspaModel)
```

## Related vignettes

- [`Multiple factors`](multiple-factors.html) — a score that is a composite of several indicators: the `fsT`/`fsL` path and `tspa()` cross-loadings.
- [`2S-PA with Random Effects`](multilevel.html) — stage 1 from an `lme4` mixed model, feeding empirical-Bayes random effects into 2S-PA.
- [`Scoring Matrices: lavaan CFA and lme4`](scoring-matrices.html) — the per-observation scoring matrices `get_fs()` attaches, and the empirical-Bayes identity behind them.
- [`Correction to Measurement Error`](correction-error.html) — the measurement-error correction factor and its effect on standard errors.
- [`EFA Scores`](efa-score.html) — stage 1 from an exploratory factor analysis (EFA) rather than a CFA.
- [`Linear Growth Modeling with 2S-PA`](tspa-growth-vignette.html) — 2S-PA for longitudinal / growth-curve measurement models.
- [`2S-PA with OpenMx and IRT (mirt)`](tspa-vignette-mx.html) — stage 2 via `tspa_mx_model()` (OpenMx), including the `mirt` (IRT) path.

## References

- Bollen, K. A. (1989). *Structural equations with latent variables*.
  New York, NY: Wiley.
- Cole, D. A., & Preacher, K. J. (2014). Manifest variable path analysis:
  Potentially serious and misleading consequences due to uncorrected measurement
  error. *Psychological Methods, 19*(2), 300–315.
  [https://doi.org/10.1037/a0033805](https://doi.org/10.1037/a0033805)
- Kline, R. B. (2016). *Principles and practice of structural equation
  modeling* (4th ed.). New York, NY: The Guilford Press.
- Lai, M. H. C., & Hsiao, Y.-Y. (2022). Two-stage path analysis with definition
  variables: An alternative framework to account for measurement error.
  *Psychological Methods, 27*(4), 568–588.
  [https://doi.org/10.1037/met0000410](https://doi.org/10.1037/met0000410)
- Lai, M. H. C., Tse, W. W.-Y., Zhang, G., Li, Y., & Hsiao, Y.-Y. (2023).
  Correcting for unreliability and partial invariance: A two-stage path analysis
  approach. *Structural Equation Modeling: A Multidisciplinary Journal, 30*(2),
  258–271.
  [https://doi.org/10.1080/10705511.2022.2125397](https://doi.org/10.1080/10705511.2022.2125397)
- Lai, M. H. C., & Liu, S. (2026). A two-stage approach to account for
  measurement error when using empirical Bayes estimates of random slopes.
  *Psychological Methods.* Advance online publication.
  [https://doi.org/10.1037/met0000838](https://doi.org/10.1037/met0000838)
- Rosseel, Y. (2012). *lavaan*: An R package for structural equation modeling.
  *Journal of Statistical Software, 48*(2), 1–36.
  [https://doi.org/10.18637/jss.v048.i02](https://doi.org/10.18637/jss.v048.i02)
