---
title: "Two-Stage Path Analysis (2S-PA) Model Examples"
author: "Yixiao Li"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{2spa_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The example is from https://lavaan.ugent.be/tutorial/sem.html. 

Load packages

```{r load-pkg, message = FALSE}
library(lavaan)
library(R2spa)
```


## Single group, single predictor^[In the model, `a`, `b`, and `c` are coefficients where variables `y1`, `y2`, `y3`, and `y4` are indicators of the latent variable `dem60`.]

```{r}
model <- ' 
  # latent variable definitions
     ind60 =~ x1 + x2 + x3
     dem60 =~ y1 + a*y2 + b*y3 + c*y4

  # regressions
    dem60 ~ ind60
'
```

To call `tspa()`, a data frame of factor scores is needed for all latent variables. To get this data frame, apply `get_fs()` to all latent variables and specify model parameter as their respective definitions. Combine factor scores for all latent variable using `cbind()` so that it can be used in `tspa()` for model building.

```{r}
fs_dat_ind60 <- get_fs(data = PoliticalDemocracy, 
                       model = "ind60 =~ x1 + x2 + x3")
fs_dat_dem60 <- get_fs(data = PoliticalDemocracy, 
                       model = "dem60 =~ y1 + y2 + y3 + y4")
fs_dat <- cbind(fs_dat_ind60, fs_dat_dem60)
```

To build a Two-Stage Path Analysis model, simply call the `tspa()` function with model = regressions, data = The combined factor score data frame using `get_fs()`, and specify standard error as either a list or a data frame. Values for standard error can be found at column named `fs_[variable name]_se`. For example, the standard error of latent variable ind60 can be found at column `fs_ind60_se` in the `fs_dat` data frame.

```{r}
tspa_fit <- tspa(model = "dem60 ~ ind60", 
                 data = fs_dat, 
                 se_fs = list(ind60 = 0.1213615, dem60 = 0.6756472))
```

To view the Two-Stage Path Analysis model, use `attributes([model name])$tspaModel`. Function `cat()` can help tidy up the model output. In the output, the values of error constraints are computed by squaring standard errors from the previous section.

```{r}
cat(attributes(tspa_fit)$tspaModel)
```

## Single group, multiple predictors

```{r}
model <- ' 
  # latent variable definitions
     ind60 =~ x1 + x2 + x3
     dem60 =~ y1 + y2 + y3 + y4
     dem65 =~ y5 + y6 + y7 + y8

  # regressions
    dem60 ~ ind60
    dem65 ~ ind60 + dem60

  # # residual correlations
  #   y1 ~~ y5
  #   y2 ~~ y4 + y6
  #   y3 ~~ y7
  #   y4 ~~ y8
  #   y6 ~~ y8
'
```

To call `tspa()`, a data frame of factor scores is needed for all latent variables. To get this data frame, apply `get_fs()` to all latent variables and specify model parameters as their respective definitions. Combine factor scores for all latent variables using `cbind()` to call `tspa()` for model building.

```{r}
fs_dat_ind60 <- get_fs(data = PoliticalDemocracy, 
                       model = "ind60 =~ x1 + x2 + x3")
fs_dat_dem60 <- get_fs(data = PoliticalDemocracy, 
                       model = "dem60 =~ y1 + y2 + y3 + y4")
fs_dat_dem65 <- get_fs(data = PoliticalDemocracy, 
                       model = "dem65 =~ y5 + y6 + y7 + y8")
fs_dat <- cbind(fs_dat_ind60, fs_dat_dem60, fs_dat_dem65)
```

To build a Two-Stage Path Analysis model, simply call the `tspa()` function with model = regressions (for all predictors), data = the factor score data frame created by combining results from `get_fs()`, and specify standard errors as either a list or a data frame. Values for standard errors can be found at column named `fs_[variable name]_se`.^[The text inside `[]` should be replaced with actual variable and model name.] For example, the standard error of latent variable ind60 can be found at column `fs_ind60_se` in the `fs_dat` data frame. In the output model, the values of error constraints are computed by squaring standard errors.

```{r}
tspa_3var_fit <- tspa(model = "dem60 ~ ind60
                          dem65 ~ ind60 + dem60", 
                      data = fs_dat, 
                      se_fs = list(ind60 = 0.1213615, dem60 = 0.6756472, 
                                   dem65 = 0.5724405))
cat(attributes(tspa_3var_fit)$tspaModel)
```

## Multigroup, single predictor

```{r}
model <- ' 
  # latent variable definitions
    visual =~ x1 + x2 + x3
    speed =~ x7 + x8 + x9

  # regressions
    visual ~ speed
'
```

To call `tspa()`, a data frame of factor scores is needed for all multigroup variables. To get this data frame, apply `get_fs()` to all multigroup variables and specify model parameter as their respective definitions. Combine factor scores for all multigroup variables using `cbind()` so that it can be fed in `tspa()` for model building.

```{r}
hs_mod <- '
visual =~ x1 + x2 + x3
speed =~ x7 + x8 + x9
'

# get factor scores
fs_dat_visual <- get_fs(data = HolzingerSwineford1939, 
                        model = "visual =~ x1 + x2 + x3", 
                        group = "school")
fs_dat_speed <- get_fs(data = HolzingerSwineford1939, 
                       model = "speed =~ x7 + x8 + x9", 
                       group = "school")
fs_hs <- cbind(fs_dat_visual, fs_dat_speed)
```

To build a Two-Stage Path Analysis model, simply call the `tspa()` function with model = regression relation. Specify standard error as either a list or a data frame. Values for standard error can be found at column named `fs_[variable name]_se`. For example, the standard error of the multigroup variable visual can be found at column `fs_visual_se` in the `fs_hs` data frame. To get the standard errors for each group faster, `unique()` can be called upon the standard error column. For example, in this case, `unique(fs_hs$fs_visual_se)` can be called to get the standard errors for multigroup variable visual.

Function `standardizedsolution()` enables user to view a table of standard error, z score, p-value, and lower bound of confidence interval for each multigroup regression relation.

```{r}
# tspa model
tspa_fit <- tspa(model = "visual ~ speed",
                 data = fs_hs,
                 se_fs = data.frame(visual = c(0.3391326, 0.311828),
                                    speed = c(0.2786875, 0.2740507)),
                 group = "school"
                 # group.equal = "regressions"
                 )
stdsol <- standardizedsolution(tspa_fit)
subset(stdsol, subset = op == "~")
```

To view the Two-Stage Path Analysis model for multigroup, use `attributes([model name])$tspaModel`. Function `cat()` can help tidy up the model output. In the output, the values of error constraints are computed by squaring standard errors from the previous section.

```{r}
cat(attributes(tspa_fit)$tspaModel)
```
