---
title: "Scoring Matrices: lavaan CFA and lme4"
author: "Mark Lai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scoring-matrices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The **scoring matrix** is the linear operator that converts the modelled
observed data into the factor / empirical-Bayes (EB) scores reported by
`get_fs()`. Together with the score error-covariance (`fsT`), implied
loadings (`fsL`) and implied intercepts (`fsb`), it fully specifies how
each reported score is constructed from the model. `get_fs()` exposes the
scoring matrix as the `scoring_matrix` attribute alongside these three:

- for **lavaan** CFA models it is a named list of one (score x item)
  matrix per group;
- for **lme4** models it is a named list of one (random-effect x
  observation) matrix per cluster.

In this vignette we reconstruct the scores **by hand** from the stored
scoring matrix and the modelled data, demonstrating the identity in both
settings. (The vignette [2S-PA with Random Effects](multilevel.html) shows
the downstream 2S-PA use of the `ev_`/`ecov_` columns that `get_fs()`
derives from `fsT`.)

```{r, message = FALSE}
library(R2spa)
library(lavaan)
library(lme4)
```

## lavaan CFA

We fit a single-factor CFA on `lavaan::PoliticalDemocracy` with a
mean structure, so that the observed-item intercepts `nu` and the factor
intercept `alpha` are estimated parameters:

```{r, message = FALSE}
cfa_fit <- cfa("f =~ y1 + y2 + y3 + y4",
               data = lavaan::PoliticalDemocracy,
               meanstructure = TRUE)
fs <- get_fs(cfa_fit)
head(fs)
```

The `scoring_matrix` attribute is a named list keyed by group label
(an empty string for single-group fits); each entry is the matrix `S`
with one row per factor and one column per item:

```{r}
sm <- attr(fs, "scoring_matrix")
S <- sm[[1]]
S
# Same results as in lavaan
attr(lavPredict(cfa_fit, fsm = TRUE), "fsm")
```

### Reconstructing the scores by hand

`get_fs()` computes the regression scores as

`f = S %*% (y' - (lambda %*% alpha + nu))' + alpha`,

where `lambda` is the item-loading matrix, `nu` the item intercepts, and
`alpha` the factor intercept. From `lavInspect(cfa_fit, "est")` we can
therefore recompute the scores directly from the raw data:

```{r}
est <- lavInspect(cfa_fit, "est")
Lambda <- as.matrix(est$lambda)
Theta  <- as.matrix(est$theta)
nu     <- as.numeric(est$nu)
alpha  <- as.numeric(est$alpha)
y      <- lavInspect(cfa_fit, "data")
fs_hand <- S %*% (t(y) - (as.vector(Lambda %*% alpha) + nu)) + as.vector(alpha)
max(abs(fs_hand - fs$fs_f))
```

The maximum deviation from the reported `fs_f` column is at the level of
machine precision. The same `S` also generates the companion attributes,
so all of them are redundant views of the same linear operator:

- `fsL = S %*% Lambda`: implied loadings of the items on the factor
  scores (i.e. the factor score's regression on the items, in loading
  form);
- `fsb = alpha - fsL %*% alpha`: implied intercepts of the factor
  scores;
- `fsT = S %*% Theta %*% t(S)`: error variance-covariance of the
  regression scores.

```{r}
max(abs(attr(fs, "fsL")[[1]] - S %*% Lambda))
max(abs(attr(fs, "fsb")[[1]] - (alpha - attr(fs, "fsL")[[1]] %*% alpha)))
max(abs(attr(fs, "fsT")[[1]] - S %*% Theta %*% t(S)))
```

## lme4

We fit the classic `sleepstudy` growth model with a random intercept and
slope for `Subject`, using ML estimation (the 2S-PA convention; see the
Notes there). Each cluster (subject) obtains a single 2-dimensional EB
estimate of the random effect:

```{r, message = FALSE}
lme1 <- lmer(Reaction ~ Days + (Days | Subject), sleepstudy,
             REML = FALSE)
fs1 <- get_fs(lme1)
head(fs1)
```

Its `scoring_matrix` attribute is a named list with one entry per
cluster; each entry `S_j` has one row per random effect and one column
per observation in the cluster:

```{r}
sm1 <- attr(fs1, "scoring_matrix")
length(sm1)
names(sm1)
knitr::kable(sm1[["308"]], digits = 3,
             caption = "Scoring matrix for subject 308")
```

### Reconstructing the scores by hand

The theoretical EB score of cluster `j` is

`u_hat_j = S_j %*% (y_j - X_j %*% beta)`,

with

`S_j = (G^{-1} + Z_j' Z_j / sigma^2)^{-1} Z_j' / sigma^2`,

where `y_j` is the cluster's response vector, `X_j` its rows of the
fixed-effects design, `Z_j` its rows of the random-effects design,
`beta = fixef(.)` the fixed-effect estimates, and `G`, `sigma^2` the
random-effects and residual variances. `R2spa` reconstructs the scaled
matrix `D = G / sigma^2` from the LME4 theta via `get_D()`; because
`Kz = Z_j' Z_j` is unscaled, `sigma^2` cancels in `S_j`. As with lavaan,
`S_j` also determines the companion attributes:

- `fsL_j = S_j %*% Z_j`: loading of the EB estimate on the random
  effect, i.e. the operator of `E[u_hat_j | u_j]`;
- `fsT_j = sigma^2 %*% S_j %*% t(S_j)`: conditional
  (measurement-error) covariance of the EB scores, `Var(u_hat_j | u_j)`.

Reconstructing the scores from the raw data per cluster:

```{r}
mf   <- model.frame(lme1)
y    <- model.response(mf)
X    <- as.matrix(lme1@pp$X)
beta <- fixef(lme1)
idx_list <- split(seq_len(nrow(mf)), lme1@flist[[1]])
fs_hand1 <- t(vapply(names(sm1), function(lv) {
  idx <- idx_list[[lv]]
  sm1[[lv]] %*% (y[idx] - as.numeric(X[idx, ] %*% beta))
}, numeric(2)))
max(abs(fs_hand1 - as.matrix(fs1[, 1:2])))
max(abs(fs_hand1 - as.matrix(ranef(lme1)[[1]])))
```

Both the `get_fs()` output and `lme4::ranef()` are recovered to
machine precision.

Under `method = "ML"` the same per-cluster identity holds with
prior-free OLS weights `S_j = (Z_j' Z_j)^+ Z_j'` (no random-effects
prior enters `S_j`), so the reconstructed scores are the per-cluster
OLS (ML) estimates of the random effects rather than the EB estimates
shown here.

### Notes

- Only the **first random-effect term** is scored (`[[1]]` convention,
  matching `ranef(lme1)[[1]]`).
- Cluster sizes may differ (unbalanced designs); the per-cluster list
  accommodates differing matrix widths.
- The fixed and random designs need not coincide (`Z != X`); the scoring
  identity is based on the random design `Z`.
- The identity holds under both ML and REML estimation; the vignette
  uses ML by 2S-PA convention.

## Comparison

```{r}
knitr::kable(
  rbind(
    c("scoring matrix `S`",
      "regression (default):
      S = Psi Lambda' (Lambda Psi Lambda' + Theta)^{-1};
      \"Bartlett\":
      S = (Lambda' Theta^{-1} Lambda)^{-1} Lambda' Theta^{-1}",
      "S_j = (G^{-1} + Z_j' Z_j / sigma^2)^{-1} Z_j' / sigma^2 (EB
      weights)"),
    c("shape & orientation",
      "score x item (q x p), shared within a group",
      "num_re x n_j per cluster j (p x n_j)"),
    c("container",
      "named list of one matrix per group",
      "named list of one matrix per cluster
      (list, not array: n_j can differ)"),
    c("score reconstruction",
      "f = S %*% (y' - (Lambda %*% alpha + nu))' + alpha
      (all cases in one product)",
      "u_hat_j = S_j %*% (y_j - X_j %*% beta)
      (one product per cluster)")
  ),
  col.names = c("property", "lavaan CFA", "lme4"),
  escape = FALSE
)
```
