---
title: "Scoring Methods and SE Correction in 2S-PA: A Simulation Study"
author: "Mark Lai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring Methods and SE Correction in 2S-PA: A Simulation Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

# Overview

Two-stage path analysis (2S-PA, [vignette("Two-Stage Path Analysis (2S-PA) Model Examples")])
estimates a structural path model by first computing factor scores (stage 1) and then
regressing those scores on one another while correcting for their measurement error
(stage 2, `tspa()`). Stage 1 can produce scores by several methods —
**regression** (empirical Bayes, the default), **Bartlett**, or **sum** scores
(`get_fs(..., method = )`). Separately, the *standard errors* of the stage-2 path
estimates can optionally be inflated by a first-order (delta-method) correction for the
sampling error in the stage-1 scoring weights
([vignette("Corrected Standard Errors")], `tspa(..., corrected_se = TRUE)`).

This vignette reports a small Monte Carlo study that places the 2S-PA point estimates
and standard errors of all three scoring methods — with and without the SE correction —
side by side with the **gold-standard joint model**, a fully joint
`lavaan::sem()` fit of the same measurement + structural model. Because the data are
simulated, the structural path has a known true value, so we can quantify bias
(point estimates) and SE accuracy (coverage).

The design:

- **Model** — the canonical two-factor path model (`ind60 =~ x1 + x2 + x3`,
  `dem60 =~ y1 + y2 + y3 + y4`, `dem60 ~ ind60`).
- **Replications** — `B = 200` datasets at each of two sample sizes, `n = 100` and
  `n = 200`.
- **Compared estimators** — for each of {regression, Bartlett, sum} scores, the 2S-PA
  path estimate with its naive SE and (where available) its corrected SE; and the
  joint-model path estimate with its SE.

# Data generation process

Each dataset is drawn from a two-factor CFA with a single structural path, on the
unit-factor-variance scale:

$$
\begin{aligned}
\texttt{ind60} & \sim N(0, 1), \\
\texttt{dem60} & = \gamma^\* \, \texttt{ind60} + \zeta, \qquad \zeta \sim N(0, 1 - (\gamma^\*)^2),\ \ \zeta \perp \texttt{ind60},
\end{aligned}
$$

so `Var(ind60) = Var(dem60) = 1` and the factor correlation is $\gamma^\* = 0.5$.
Indicators load on their own factor with errors independent of the factor:

```{r dgp, message = FALSE}
library(lavaan)
library(R2spa)
x <- readRDS("sim_scoring_se.RDS")
meta <- x$meta
c(
  "gamma* (true std.all path) =", meta$gamma_star,
  "ind60 loadings (x1,x2,x3)  =", paste(round(meta$lam_ind, 2), collapse = ", "),
  "dem60 loadings (y1..y4)    =", paste(round(meta$lam_dem, 2), collapse = ", "),
  "n =", paste(meta$n, collapse = ", "), "  B =", meta$B, "  seed =", meta$seed
)
```

The true quantity of interest is the **standardized** (std.all) structural path, which
equals $\gamma^\* = 0.5$ on this scale.

# Estimators compared

## Gold-standard joint model

The reference is the fully joint `lavaan::sem()` of the measurement model and the
structural path, fit with **marker indicators** (not `std.lv`). We read the structural
path on the **standardized (std.all) scale** via `standardizedSolution()`.

> **Why not `std.lv = TRUE`?** In a *structural* model `std.lv` fixes each latent's
> variance to 1 by pinning its `~~` parameter. For an exogenous latent that `~~` is the
> latent variance, but for the endogenous `dem60` it is the **residual** variance, so
> `std.lv` mis-fixes `dem60 ~~ dem60` to 1 and biases the path (we verified it drifts
> to ~0.58 here). Marker indicators avoid this; `standardizedSolution()` then puts the
> path on the scale where the true value is $\gamma^\* = 0.5$.

## 2S-PA

For each scoring method `m` in {`"regression"`, `"Bartlett"`, `"mean"`}:

1. Stage 1 — fit the CFA (`std.lv = TRUE`, the usual scoring scale) and score it:
   `get_fs(fit, method = m, vfsLT = TRUE)` (sum scores omit `vfsLT`, which they do not
   support).
2. Stage 2 — `tspa("dem60 ~ ind60", scores)`, reading `fsL`/`fsT` from the stage-1
   result. The naive SE is `vcov()` of that fit; the corrected SE additionally passes
   `corrected_se = TRUE` with the stage-1 `vfsLT`.
3. Read the path and its SE on the std.all scale via `standardizedSolution()`.

Two scoring-method details shape the corrected SE:

- **Bartlett** scores fix the loadings (identity `fsL`), so only the score error
  variances are free stage-1 quantities; the correction propagates those only
  (`which_free` = the free `fsT` positions).
- **regression** scores have free loadings *and* error variances, so the correction
  propagates more stage-1 uncertainty.
- **sum** scores support neither `vfsLT` nor the correction, so they enter the SE
  comparison with the naive SE only.

# Results

```{r summ, message = FALSE}
res <- x$res
gamma_star <- meta$gamma_star

summ <- function(g, se) {
  lo <- g - 1.96 * se; hi <- g + 1.96 * se
  c(bias = mean(g) - gamma_star, mc_sd = sd(g),
    mean_se = mean(se), calib = mean(se) / sd(g),
    coverage = mean(lo <= gamma_star & gamma_star <= hi))
}
cov95 <- function(g, se) mean((g - 1.96 * se) <= gamma_star &
                                  (g + 1.96 * se) >= gamma_star)
n_list <- meta$n
```

## Point estimates

The bias and Monte Carlo SD of the std.all path estimate, per method and sample size
(the corrected SE does not change the point estimate, so it is shown once per method):

```{r point-est}
pt <- do.call(rbind, lapply(n_list, function(n) {
  d <- res[res$n == n, ]
  data.frame(
    n = n,
    method = c("regression", "Bartlett", "sum", "joint (gold)"),
    bias = c(mean(d$est_reg) - gamma_star, mean(d$est_bar) - gamma_star,
             mean(d$est_sum) - gamma_star, mean(d$est_joint) - gamma_star),
    mc_sd = c(sd(d$est_reg), sd(d$est_bar), sd(d$est_sum), sd(d$est_joint)),
    stringsAsFactors = FALSE)
}))
pt[, c("bias", "mc_sd")] <- round(pt[, c("bias", "mc_sd")], 4)
knitr::kable(pt, row.names = FALSE,
             caption = "Point estimates (std.all path). True value = 0.5.")
```

All scoring methods — and the joint model — are essentially unbiased and have the same
Monte Carlo SD. 2S-PA recovers the gold-standard point estimate for every scoring
method; the choice of scoring method does not bias the structural estimate.

## Standard errors

The mean reported SE, its calibration (mean SE / Monte Carlo SD), and the empirical
95% coverage, per method and SE treatment:

```{r se-table}
se_tbl <- do.call(rbind, lapply(n_list, function(n) {
  d <- res[res$n == n, ]
  data.frame(
    n = n,
    method = c("regression", "regression", "Bartlett", "Bartlett", "sum", "joint (gold)"),
    se = c("naive", "corrected", "naive", "corrected", "naive", "naive"),
    mean_se = c(mean(d$se_reg_naive), mean(d$se_reg_corr),
                mean(d$se_bar_naive), mean(d$se_bar_corr),
                mean(d$se_sum_naive), mean(d$se_joint)),
    calib = c(mean(d$se_reg_naive) / sd(d$est_reg), NA,
              mean(d$se_bar_naive) / sd(d$est_bar), NA,
              mean(d$se_sum_naive) / sd(d$est_sum),
              mean(d$se_joint) / sd(d$est_joint)),
    coverage = c(summ(d$est_reg, d$se_reg_naive)["coverage"],
                 cov95(d$est_reg, d$se_reg_corr),
                 summ(d$est_bar, d$se_bar_naive)["coverage"],
                 cov95(d$est_bar, d$se_bar_corr),
                 summ(d$est_sum, d$se_sum_naive)["coverage"],
                 summ(d$est_joint, d$se_joint)["coverage"]),
    stringsAsFactors = FALSE)
}))
se_tbl[, c("mean_se", "calib", "coverage")] <-
  round(se_tbl[, c("mean_se", "calib", "coverage")], 3)
knitr::kable(se_tbl, row.names = FALSE,
             caption = "Standard errors (std.all path). True value = 0.5; nominal coverage 0.95.")
```

## Coverage

```{r coverage-plot, message = FALSE, fig.width = 8, fig.height = 4}
covdat <- do.call(rbind, lapply(n_list, function(n) {
  d <- res[res$n == n, ]
  data.frame(
    n = n,
    label = c("regression naive", "regression corr.", "Bartlett naive",
              "Bartlett corr.", "sum naive", "joint"),
    se = c("naive", "corrected", "naive", "corrected", "naive", "naive"),
    coverage = c(summ(d$est_reg, d$se_reg_naive)["coverage"],
                 cov95(d$est_reg, d$se_reg_corr),
                 summ(d$est_bar, d$se_bar_naive)["coverage"],
                 cov95(d$est_bar, d$se_bar_corr),
                 summ(d$est_sum, d$se_sum_naive)["coverage"],
                 summ(d$est_joint, d$se_joint)["coverage"]))
}))
op <- par(mfrow = c(1, 2), mar = c(8, 4, 3, 1))
for (n in n_list) {
  sub <- covdat[covdat$n == n, ]
  cols <- ifelse(sub$se == "corrected", "#2166ac", "grey70")
  barplot(sub$coverage, names.arg = sub$label, col = cols,
          ylim = c(0.85, 1.0), las = 2, cex.names = 0.6,
          main = sprintf("n = %d", n), ylab = "empirical 95% coverage")
  abline(h = 0.95, lty = 2, col = "grey30")
}
par(op)
```

# Discussion

- **Point estimates track the gold standard.** Every scoring method is (near-)
  unbiased and matches the joint model's point estimate and precision. The choice
  among regression, Bartlett, and sum scoring does not bias the structural estimate.
- **Naive SEs already track the joint model.** The naive 2S-PA SE is close to the
  joint model's SE at both sample sizes (both are slightly anti-conservative at
  `n = 100`, a familiar small-sample finite-sample effect of likelihood-based SEs).
- **The SE correction adds conservatism.** Inflating the SE for the sampling error in
  the stage-1 weights pushes coverage above 95%. The effect is much larger for
  **regression** scoring (free loadings and error variances are propagated) than for
  **Bartlett** (only the error variances are free), and it shrinks as `n` grows.
  **Sum** scores cannot use the correction (no `vfsLT`).

The practical takeaway: for a well-specified model of this size, all three scoring
methods recover the joint-model point estimate, and the naive 2S-PA SE is already close
to the joint-model SE. The delta-method SE correction is a conservative refinement whose
impact depends on how much stage-1 uncertainty the scoring method leaves free
(regression > Bartlett) and on the sample size.

# Reproducibility

The per-replication table is cached in `sim_scoring_se.RDS` (this vignette's directory);
the code below regenerates it. It runs in a few minutes.

```{r sim-loop, eval = FALSE}
gamma_star <- 0.5
lam_ind <- c(0.80, 0.75, 0.85)   # ind60 =~ x1 + x2 + x3
lam_dem <- c(0.75, 0.80, 0.70, 0.85)   # dem60 =~ y1 + y2 + y3 + y4
cfa_model <- "ind60 =~ x1 + x2 + x3\ndem60 =~ y1 + y2 + y3 + y4"
sem_model <- paste0(cfa_model, "\ndem60 ~ ind60")   # marker indicators, no std.lv

sim_one <- function(n) {
  ind <- rnorm(n)
  dem <- gamma_star * ind + rnorm(n, sd = sqrt(1 - gamma_star^2))
  data.frame(
    x1 = lam_ind[1] * ind + rnorm(n, sd = sqrt(1 - lam_ind[1]^2)),
    x2 = lam_ind[2] * ind + rnorm(n, sd = sqrt(1 - lam_ind[2]^2)),
    x3 = lam_ind[3] * ind + rnorm(n, sd = sqrt(1 - lam_ind[3]^2)),
    y1 = lam_dem[1] * dem + rnorm(n, sd = sqrt(1 - lam_dem[1]^2)),
    y2 = lam_dem[2] * dem + rnorm(n, sd = sqrt(1 - lam_dem[2]^2)),
    y3 = lam_dem[3] * dem + rnorm(n, sd = sqrt(1 - lam_dem[3]^2)),
    y4 = lam_dem[4] * dem + rnorm(n, sd = sqrt(1 - lam_dem[4]^2)))
}

stdall <- function(fit) {
  s <- standardizedSolution(fit, type = "std.all")
  i <- which(s$lhs == "dem60" && s$op == "~" && s$rhs == "ind60")
  c(est = s$est.std[i], se = s$se[i])
}
corr_fit <- function(fs) {
  v <- attr(fs, "vfsLT"); wf <- which(diag(v) > 1e-12)
  tspa("dem60 ~ ind60", fs, vfsLT = v[wf, wf], which_free = wf,
       corrected_se = TRUE)
}

run_rep <- function(n) {
  d   <- sim_one(n)
  cf  <- cfa(cfa_model, d, std.lv = TRUE)
  sem <- sem(sem_model, d)
  sr  <- get_fs(cf, method = "regression", vfsLT = TRUE)
  sb  <- get_fs(cf, method = "Bartlett",  vfsLT = TRUE)
  ss  <- get_fs(cf, method = "mean")
  tr  <- tspa("dem60 ~ ind60", sr)
  tb  <- tspa("dem60 ~ ind60", sb)
  ts  <- tspa("dem60 ~ ind60", ss)
  trc <- corr_fit(sr); tbc <- corr_fit(sb)
  g <- stdall(sem); gr <- stdall(tr); gsc <- stdall(trc)
  gb <- stdall(tb); gscb <- stdall(tbc); gs <- stdall(ts)
  c(est_reg = gr[["est"]], se_reg_naive = gr[["se"]], se_reg_corr = gsc[["se"]],
    est_bar = gb[["est"]], se_bar_naive = gb[["se"]], se_bar_corr = gscb[["se"]],
    est_sum = gs[["est"]], se_sum_naive = gs[["se"]],
    est_joint = g[["est"]], se_joint = g[["se"]])
}

n_list <- c(100, 200); B <- 200
set.seed(20260830)
res <- do.call(rbind, lapply(n_list, function(n) {
  out <- as.data.frame(t(vapply(seq_len(B), function(b) run_rep(n), numeric(10))))
  out$n <- n; out$rep <- seq_len(B); out
}))
saveRDS(list(res = res, meta = list(gamma_star = gamma_star,
                                    lam_ind = lam_ind, lam_dem = lam_dem,
                                    n = n_list, B = B, seed = 20260830)),
        file = "sim_scoring_se.RDS")
```
