---
title: "2S-PA with Categorical Items (mirt): A Simulation Study"
author: "Mark Lai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2S-PA with Categorical Items (mirt): A Simulation Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

Two-stage path analysis (2S-PA) estimates a structural path model by first computing
factor scores (stage 1) and then regressing those scores on one another while
correcting for their measurement error (stage 2, `tspa()` / `tspa_mx_model()`).
Until now the package's scoring and correction machinery was exercised on *continuous*
indicators. This vignette reports a small Monte Carlo study that asks whether the
two-stage approximation remains accurate when the stage-1 model is a **categorical**
(item-response) fit scored by `get_fs.mirt()`, and how it compares with the
fully joint **WLSMV** model.

The structural model has two correlated unit-variance factors with a single path,
`F2 ~ F1`. The stage-1 items are either **binary** (2PL) or **polytomous** (graded
response, 4 categories), and the item thresholds are either **balanced** (spread
across the latent range) or **skewed** (shifted so most responses fall in the lowest
category). The estimators differ along two dimensions — the **stage-1 scoring** (joint
vs. local per-construct) and the **stage-2 correction** (pooled vs. exact):

- **`tspa`** / **`tspa_mx`** — two-stage path analysis with the *joint* stage 1 and,
  respectively, the pooled lavaan or the exact OpenMx stage 2;
- **`tspa_local`** / **`tspa_mx_local`** — the same two stage-2 routes, but with a
  *local* stage 1 that scores each factor from its own items separately;
- **WLSMV** — the fully joint `lavaan::sem()` with categorical indicators (gold standard).

Because the data are simulated, the true standardized path is known, so we can report
bias (point estimate) and coverage (standard-error accuracy).

The generation script is `data-raw/sim_categorical_irt.R` (kept out of the build); it
writes the fixture read here (`vignettes/sim_categorical_irt.RDS`).

# Data generation process

The latent model is, on the unit-factor-variance scale,

$$
\begin{aligned}
\texttt{F1} & \sim N(0, 1), \\
\texttt{F2} & = 0.6 \, \texttt{F1} + \zeta, \qquad \zeta \sim N(0, 0.64),
\end{aligned}
$$

so `Var(F1) = Var(F2) = 1`, the factor correlation is $0.6$, and the true
**standardized** `F2 ~ F1` path is $\gamma^* = 0.6$. Each factor is measured by
`5` items with discrimination $a = 1.5$.

- **Binary** items follow the 2PL, $P(Y = 1 \mid \theta) = 1 / (1 + e^{-(a\theta - b)})$.
- **Polytomous** items follow the graded-response model with `4` categories and
  thresholds $d_1 < d_2 < d_3$, where $P(Y \ge k \mid \theta) = 1 / (1 + e^{-(a\theta - d_k)})$.

The two threshold conditions shift the marginal item response:

```{r dgp, message = FALSE}
x <- readRDS("sim_categorical_irt.RDS")
meta <- x$meta
cat("gamma* (true std path) =", meta$gamma_star,
    "  n =", paste(meta$n, collapse = ", "),
    "  B =", meta$B, "\n")
cat("mean marginal item response by condition:\n")
print(round(meta$marginal, 3))
```

`balanced` is centered (binary marginal ~0.5; graded ~1.4 of 0--3) while `skewed`
concentrates responses in the low end (binary ~0.2; graded ~0.7). Skewed thresholds
make many observations weakly identified, so the per-observation score standard errors
are larger and more heterogeneous — the condition that stresses the two-stage
correction.

# Estimators compared

All five estimate the **standardized** `F2 ~ F1` path (the correlation between the
factors), the scale-free quantity comparable across them. They differ along two
dimensions — the **stage-1 scoring** (joint vs. local) and the **stage-2 correction**
(pooled vs. exact):

- **Stage 1 — joint.** Fit the two-factor item-response model (`mirt`, `COV = F1*F2`) and
  score it with `get_fs()`. Each factor's EAP score conditions on *all* items, so the two
  factors borrow information from one another through their correlation.
- **Stage 1 — local.** Score each factor from its *own* five items with two separate
  single-factor `mirt` fits, then combine the two single-factor results into a
  block-diagonal two-factor result with `combine_fs()` (zero cross-block `fsL`/`fsT`).
  Each score then conditions only on its own items.
- **Stage 2 — pooled** (`tspa`, `tspa_local`): `tspa("F2 ~ F1", scores)`, which *pools*
  the per-observation quantities to a representative value before correcting the
  regression.
- **Stage 2 — exact** (`tspa_mx`, `tspa_mx_local`): `tspa_mx_model("F2 ~ F1", scores)`,
  which uses the per-observation quantities as definition variables — the non-pooled
  correction.
- **WLSMV:** the fully joint `lavaan::sem()` of the measurement model and the path, fit
  with `estimator = "WLSMV"` on the ordered items. It depends only on the raw items, so
  it is the same reference regardless of stage-1 scoring.

The standardized path is read from each fitted model as
$\hat\gamma \, \text{SD}(F1) / \text{SD}(F2)$, with the standard error from a
first-order (delta-method) linearization in the structural path and the two factor
variances. This single formula reproduces `lavaan`'s `standardizedSolution()` for the
two lavaan arms and the equivalent algebra for the OpenMx arm.

# Results

`B = `r meta$B` replications at each of the `r length(meta$n)` sample sizes
(`n = `r paste(meta$n, collapse = ", ")`)). A finite point estimate was obtained in at
least `92%` of replications in every condition, and in at least `99%` at `n = 200` and
`500`. A few `n = 50` fits returned no usable standard error; those reps are excluded
from the SE and coverage statistics only (their point estimates still count toward bias).

```{r summ, message = FALSE}
res <- x$res
g <- meta$gamma_star
summ <- function(d) {
  de <- d[!is.na(d$est_std), ]
  ds <- d[!is.na(d$se_std), ]
  lo <- ds$est_std - 1.96 * ds$se_std; hi <- ds$est_std + 1.96 * ds$se_std
  c(bias = mean(de$est_std) - g, mc_sd = sd(de$est_std),
    rmse = sqrt(mean((de$est_std - g)^2)),
    mean_se = mean(ds$se_std), calib = mean(ds$se_std) / sd(de$est_std),
    coverage = mean(lo <= g & g <= hi), n_se = nrow(ds))
}
methods <- c("tspa", "tspa_mx", "WLSMV", "tspa_local", "tspa_mx_local")
```

## Point estimates

Per condition and method, bias (mean estimate minus `0.6`), the Monte Carlo standard
deviation (MC SD), and RMSE (`sqrt(bias^2 + MC SD^2)`) of the standardized path, with the
binary and polytomous items in separate columns:

```{r point-est}
pt_ns <- expand.grid(n = meta$n, skew = c("balanced", "skewed"),
                     stringsAsFactors = FALSE)
pt <- do.call(rbind, lapply(seq_len(nrow(pt_ns)), function(i) {
  o <- pt_ns[i, ]
  mm <- function(it, m) {
    d <- res[res$n == o$n & res$itemtype == it & res$skew == o$skew & res$method == m, ]
    s <- summ(d); c(s["bias"], s["mc_sd"], s["rmse"])
  }
  do.call(rbind, lapply(methods, function(m) {
    b <- mm("binary", m); gd <- mm("graded", m)
    data.frame(n = o$n, skew = o$skew, method = m,
               bias_binary = round(b[1], 3), mc_sd_binary = round(b[2], 3),
               rmse_binary = round(b[3], 3),
               bias_graded = round(gd[1], 3), mc_sd_graded = round(gd[2], 3),
               rmse_graded = round(gd[3], 3), stringsAsFactors = FALSE)
  }))
}))
knitr::kable(pt, row.names = FALSE,
             caption = "Point estimates (standardized path) by item type. True value = 0.6.")
```

At `n = 200` and `500` the joint methods (`tspa`, `tspa_mx`) and WLSMV are close to
unbiased (cell biases within about `0.02` of `0`). The **pooled** `tspa` is biased
downward in every one of those cells, and the **exact** `tspa_mx` has strictly smaller
bias than the pooled route; the WLSMV reference has the smallest bias in most cells. The pooled-vs-exact gap is largest under
the **skewed** thresholds, where the per-observation standard errors are most
heterogeneous. The smaller sample size (`n = 50`) is noisier, as expected (Monte Carlo SD
roughly two to three times larger), but the point estimates stay close to `0.6` in most
cells. The one clear exception is the WLSMV reference on the **skewed binary** items,
which is biased upward (mean estimate about `0.70`, bias about `+0.10`) and has the worst
coverage in the study — consistent with WLSMV's sensitivity to the low-frequency
categories that skewed binary thresholds produce at small `n`.

## Standard errors

Mean reported standard error, its calibration (`mean SE` / Monte Carlo SD), and the
empirical 95% coverage, per condition and method, with the binary and polytomous items in
separate columns (`n_se` = reps with a usable SE):

```{r se-table}
se <- do.call(rbind, lapply(seq_len(nrow(pt_ns)), function(i) {
  o <- pt_ns[i, ]
  mm <- function(it, m) {
    d <- res[res$n == o$n & res$itemtype == it & res$skew == o$skew & res$method == m, ]
    s <- summ(d); c(s["mean_se"], s["calib"], s["coverage"], s["n_se"])
  }
  do.call(rbind, lapply(methods, function(m) {
    b <- mm("binary", m); gd <- mm("graded", m)
    data.frame(n = o$n, skew = o$skew, method = m,
               mean_se_binary = round(b[1], 3), calib_binary = round(b[2], 3),
               coverage_binary = round(b[3], 3), n_se_binary = b[4],
               mean_se_graded = round(gd[1], 3), calib_graded = round(gd[2], 3),
               coverage_graded = round(gd[3], 3), n_se_graded = gd[4],
               stringsAsFactors = FALSE)
  }))
}))
knitr::kable(se, row.names = FALSE,
             caption = "Standard errors (standardized path) by item type. Nominal coverage 0.95.")
```

At `n = 200` and `500` coverage is within about `4` points of the nominal `0.95`, and
calibration is condition-dependent rather than uniformly close to `1` (near- or
above-nominal for the balanced binary conditions, mildly below for the graded and skewed
items). At `n = 50` the picture changes: the first-order (delta-method) SEs used throughout
**under-cover** for all three methods (calibration roughly `0.7` to `0.95`), because a
first-order linearization of the standardized path understates its sampling variability when
the stage-2 parameters are themselves uncertain. The WLSMV SEs are the least calibrated in
these small-sample cells (about `0.7`), and on the skewed binary items the usable-SE count
(`n_se`) also drops, since some fits return no usable covariance.

## Joint vs. local stage-1 scoring

The `tspa_local` / `tspa_mx_local` arms repeat the two-stage estimation with *local*
(per-construct) scoring. Local and joint scoring give nearly the same precision (Monte
Carlo SD) and similar standard-error coverage at every `n` — the stage-2 correction
absorbs the difference in how the scores were obtained. The difference is in **bias at
small `n`**: at `n = 50` the locally-scored arms are noticeably more biased (mean cell
`|bias|` about `0.10`, versus about `0.03`–`0.04` for the joint arms), the penalty
largest for the **binary** items, where each item carries less information and dropping
the cross-factor borrowing hurts most. At `n = 200` and `500` the gap closes to a couple
of hundredths or less, so the two scoring modes are essentially equivalent at moderate
and larger sample sizes. Per-construct scoring is therefore a workable alternative to
joint scoring, at the cost of some extra small-sample bias.

# Discussion

- **The two-stage approximation carries over to categorical items.** At `n = 200` and `500`
  both `tspa` and `tspa_mx` recover the standardized path for binary and polytomous
  indicators, matching the WLSMV reference to within about `0.03` in the worst cell (the
  exact `tspa_mx` route within about `0.015`). Categorical stage-1 items are therefore not
  an obstacle to 2S-PA.
- **Exact beats pooled under skew.** Pooling the per-observation error variances to a
  single representative value is a better approximation when those variances are
  homogeneous (balanced thresholds) and a worse one when they are heterogeneous
  (skewed thresholds, where many observations are weakly identified). The bias gap
  between `tspa` and `tspa_mx` is largest in exactly those cells; the OpenMx route, which
  keeps the per-observation quantities, stays closer to WLSMV.
- **SEs are well behaved at moderate `n`, with their limits visible at `n = 50`.** At
  `n = 200` and `500` all three methods give coverage within about `4` points of `0.95`
  with near-nominal, condition-dependent calibration. At `n = 50` the first-order
  delta-method SEs under-cover for every method (calibration about `0.7`–`0.95`) — an
   expected limitation of first-order linearization when the stage-2 parameters are
   themselves uncertain — and WLSMV is the least calibrated (about `0.7`), with the fewest
   usable SEs on the skewed binary items.
- **Local (per-construct) scoring is close to joint, except at small `n`.** Scoring each
  factor from its own items and combining the results block-diagonally with
  `combine_fs()` gives essentially the same estimates as joint scoring at `n = 200` and
  `500`, with similar precision and SE coverage at every size. The only cost is extra bias
  at `n = 50` (largest for binary items): joint EAP scoring's cross-factor borrowing is
  worth more when each factor is measured by few, weak items.
- **Scope.** The study is small-scale (`B = `r meta$B`), single-group, and uses one
  discrimination value and `4` categories for the polytomous case. It is meant to show
  that the machinery works and to point at the pooled-vs-exact trade-off, not to give
  precise coverage figures.
