---
title: "2S-PA with Categorical Items (mirt): A Simulation Study"
author: "Mark Lai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2S-PA with Categorical Items (mirt): A Simulation Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

Two-stage path analysis (2S-PA) estimates a structural path model by first computing
factor scores (stage 1) and then regressing those scores on one another while
correcting for their measurement error (stage 2, `tspa()` / `tspa_mx_model()`).
Until now the package's scoring and correction machinery was exercised on *continuous*
indicators. This vignette reports a small Monte Carlo study that asks whether the
two-stage approximation remains accurate when the stage-1 model is a **categorical**
(item-response) fit scored by `get_fs.mirt()`, and how it compares with the
fully joint **WLSMV** model.

The structural model has two correlated unit-variance factors with a single path,
`F2 ~ F1`. The stage-1 items are either **binary** (2PL) or **polytomous** (graded
response, 4 categories), and the item thresholds are either **balanced** (spread
across the latent range) or **skewed** (shifted so most responses fall in the lowest
category). The three estimators compared are:

- **`tspa`** — two-stage path analysis with the pooled lavaan stage 2;
- **`tspa_mx`** — two-stage path analysis with the exact OpenMx stage 2;
- **WLSMV** — the fully joint `lavaan::sem()` with categorical indicators (gold standard).

Because the data are simulated, the true standardized path is known, so we can report
bias (point estimate) and coverage (standard-error accuracy).

The generation script is `data-raw/sim_categorical_irt.R` (kept out of the build); it
writes the fixture read here (`vignettes/sim_categorical_irt.RDS`).

# Data generation process

The latent model is, on the unit-factor-variance scale,

$$
\begin{aligned}
\texttt{F1} & \sim N(0, 1), \\
\texttt{F2} & = 0.6 \, \texttt{F1} + \zeta, \qquad \zeta \sim N(0, 0.64),
\end{aligned}
$$

so `Var(F1) = Var(F2) = 1`, the factor correlation is $0.6$, and the true
**standardized** `F2 ~ F1` path is $\gamma^* = 0.6$. Each factor is measured by
`5` items with discrimination $a = 1.5$.

- **Binary** items follow the 2PL, $P(Y = 1 \mid \theta) = 1 / (1 + e^{-(a\theta - b)})$.
- **Polytomous** items follow the graded-response model with `4` categories and
  thresholds $d_1 < d_2 < d_3$, where $P(Y \ge k \mid \theta) = 1 / (1 + e^{-(a\theta - d_k)})$.

The two threshold conditions shift the marginal item response:

```{r dgp, message = FALSE}
x <- readRDS("sim_categorical_irt.RDS")
meta <- x$meta
cat("gamma* (true std path) =", meta$gamma_star,
    "  n =", paste(meta$n, collapse = ", "),
    "  B =", meta$B, "\n")
cat("mean marginal item response by condition:\n")
print(round(meta$marginal, 3))
```

`balanced` is centered (binary marginal ~0.5; graded ~1.4 of 0--3) while `skewed`
concentrates responses in the low end (binary ~0.2; graded ~0.7). Skewed thresholds
make many observations weakly identified, so the per-observation score standard errors
are larger and more heterogeneous — the condition that stresses the two-stage
correction.

# Estimators compared

All three estimate the **standardized** `F2 ~ F1` path (the correlation between the
factors), which is the scale-free quantity comparable across them.

- **Stage 1 (shared by the two 2S-PA arms):** fit the two-factor item-response model
  (`mirt`, `COV = F1*F2`) and score it with `get_fs()`. This returns the EAP scores and
  the per-observation error-variance quantities (`fsT`/`fsL`/`fsb`) that drive the
  correction.
- **`tspa`:** stage 2 is `tspa("F2 ~ F1", scores)`, which *pools* the per-observation
  quantities to a representative value before correcting the stage-2 regression.
- **`tspa_mx`:** stage 2 is `tspa_mx_model("F2 ~ F1", scores)`, which uses the
  per-observation quantities as definition variables — the *exact*, non-pooled
  correction.
- **WLSMV:** the fully joint `lavaan::sem()` of the measurement model and the path, fit
  with `estimator = "WLSMV"` on the ordered items.

The standardized path is read from each fitted model as
$\hat\gamma \, \text{SD}(F1) / \text{SD}(F2)$, with the standard error from a
first-order (delta-method) linearization in the structural path and the two factor
variances. This single formula reproduces `lavaan`'s `standardizedSolution()` for the
two lavaan arms and the equivalent algebra for the OpenMx arm.

# Results

`B = `r meta$B` replications at each of two sample sizes. Convergence was at least
`99%` of replications in every condition for every method.

```{r summ, message = FALSE}
res <- x$res
g <- meta$gamma_star
summ <- function(d) {
  d <- d[!is.na(d$est_std), ]
  lo <- d$est_std - 1.96 * d$se_std; hi <- d$est_std + 1.96 * d$se_std
  c(bias = mean(d$est_std) - g, mc_sd = sd(d$est_std),
    mean_se = mean(d$se_std), calib = mean(d$se_std) / sd(d$est_std),
    coverage = mean(lo <= g & g <= hi))
}
conds <- expand.grid(n = meta$n, item = c("binary", "graded"),
                     skew = c("balanced", "skewed"), stringsAsFactors = FALSE)
methods <- c("tspa", "tspa_mx", "WLSMV")
```

## Point estimates

Bias (mean estimate minus `0.6`) and the Monte Carlo standard deviation of the
standardized path, per condition and method:

```{r point-est}
pt <- do.call(rbind, lapply(seq_len(nrow(conds)), function(i) {
  o <- conds[i, ]
  do.call(rbind, lapply(methods, function(m) {
    d <- res[res$n == o$n & res$itemtype == o$item & res$skew == o$skew & res$method == m, ]
    s <- summ(d)
    data.frame(n = o$n, item = o$item, skew = o$skew, method = m,
               bias = round(s["bias"], 3), mc_sd = round(s["mc_sd"], 3),
               stringsAsFactors = FALSE)
  }))
}))
knitr::kable(pt, row.names = FALSE,
             caption = "Point estimates (standardized path). True value = 0.6.")
```

All three methods are close to unbiased (cell biases within about `0.02` of `0.6`). The
**pooled** `tspa` is biased downward in every cell, and the **exact** `tspa_mx` has
strictly smaller bias than the pooled route in every cell; the WLSMV reference has the
smallest bias in most cells. The pooled-vs-exact gap is largest under the **skewed**
thresholds — largest in the skewed binary, `n = 200`, cell — where the per-observation
standard errors are most heterogeneous.

## Standard errors

Mean reported standard error, its calibration (`mean SE` / Monte Carlo SD), and the
empirical 95% coverage, per condition and method:

```{r se-table}
se <- do.call(rbind, lapply(seq_len(nrow(conds)), function(i) {
  o <- conds[i, ]
  do.call(rbind, lapply(methods, function(m) {
    d <- res[res$n == o$n & res$itemtype == o$item & res$skew == o$skew & res$method == m, ]
    s <- summ(d)
    data.frame(n = o$n, item = o$item, skew = o$skew, method = m,
               mean_se = round(s["mean_se"], 3), calib = round(s["calib"], 3),
               coverage = round(s["coverage"], 3), stringsAsFactors = FALSE)
  }))
}))
knitr::kable(se, row.names = FALSE,
             caption = "Standard errors (standardized path). Nominal coverage 0.95.")
```

Coverage is within about `4` points of the nominal `0.95` everywhere. Calibration is
condition-dependent rather than uniformly close to `1`: it sits near or above `1` for the
balanced binary conditions, falls below `1` for the graded items (most so at `n = 500`),
and it does **not** improve monotonically with `n` — for the balanced graded and skewed
binary conditions the reported SE is, if anything, *more* conservative (further below `1`)
as `n` grows.

# Discussion

- **The two-stage approximation carries over to categorical items.** Both `tspa` and
  `tspa_mx` recover the standardized path for binary and polytomous indicators, matching
  the WLSMV reference to within about `0.03` in the worst (skewed, `n = 200`) cell — the
  exact `tspa_mx` route stays within about `0.015` throughout. Categorical stage-1 items
  are therefore not an obstacle to 2S-PA.
- **Exact beats pooled under skew.** Pooling the per-observation error variances to a
  single representative value is a better approximation when those variances are
  homogeneous (balanced thresholds) and a worse one when they are heterogeneous
  (skewed thresholds, where many observations are weakly identified). The bias gap
  between `tspa` and `tspa_mx` is largest in exactly those cells; the OpenMx route, which
  keeps the per-observation quantities, stays closer to WLSMV.
- **SEs are well behaved.** All three methods give coverage within about `4` points of
  `0.95`. Calibration is condition-dependent — near- or above-nominal for balanced binary,
  mildly conservative for the graded and skewed items — and does not improve uniformly with
  `n`, but the deviations from nominal are small.
- **Scope.** The study is small-scale (`B = `r meta$B`), single-group, and uses one
  discrimination value and `4` categories for the polytomous case. It is meant to show
  that the machinery works and to point at the pooled-vs-exact trade-off, not to give
  precise coverage figures.
