---
title: "2S-PA with a brms location-scale model (random sigma)"
author: "Mark Lai"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{2S-PA with a brms location-scale model (random sigma)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Two-stage path analysis (2S-PA; Lai & Hsiao, 2022) estimates a structural model
whose constructs are measured with error in two steps. **Stage 1** scores each
construct and records its measurement error; **stage 2** path-analyzes the
scores while correcting for that error, so the structural coefficients are not
biased by it (Cole & Preacher, 2014). In `R2spa`, `get_fs()` is stage 1 and
`tspa()` is stage 2.

This vignette uses stage 1 from a **brms location-scale mixed model** — a
Gaussian model in which the residual (measurement) scale has its own random
effect, a *random sigma*. A random sigma makes the measurement error
**person-specific** (heteroskedastic): subjects with a larger residual variance
have noisier scores. `get_fs()` recovers each subject's score *and* its
person-specific standard error.

To make the comparison concrete, we simulate data with a **known** structural
coefficient, fit the brms location-scale model (and a small CFA for a distal
construct), and regress the distal construct on the trait in three ways at stage
2: a naive regression that ignores the measurement error, the standard `tspa()`
(a `lavaan` model that pools the person-specific error into one value), and
`tspa_mx_model()` (an `OpenMx` model that uses the error **per subject**, via
definition variables — the exact, non-pooled route).

## The data

We simulate `n` subjects, each nested within one of `n_sites` sites, with `m`
noisy measurements per subject. Each subject has a latent trait (the random
intercept) and a person-specific noise level (the random sigma); the measured
values are the trait plus person-specific noise. A **distal construct** (one per
subject) is generated from the subject's *true* trait with a known coefficient
`beta1` and is measured by three noisy indicators, so it is itself scored as a
latent.

```{r dgp, eval = FALSE}
suppressMessages({ library(brms); library(R2spa) })
set.seed(33)

n_sites <- 20; per_site <- 10; n <- n_sites * per_site; m <- 5
beta1 <- 1.0
p      <- factor(rep(1:n_sites, each = per_site))  # site (higher level)
b0     <- rnorm(n, 0, 0.5)                         # true latent trait
sig    <- exp(rnorm(n, -1.0, 0.5))                 # person-specific noise SD
sid    <- factor(rep(1:n, each = m))
y      <- b0[as.integer(sid)] + sig[as.integer(sid)] * rnorm(n * m)
d      <- data.frame(y = y, Subject = sid, p = p[as.integer(sid)])
# distal construct, measured by three noisy indicators
d0     <- beta1 * b0 + rnorm(n, 0, 0.5)
dj     <- data.frame(d1 = d0 + rnorm(n, 0, 0.5),
                     d2 = d0 + rnorm(n, 0, 0.5),
                     d3 = d0 + rnorm(n, 0, 0.5))
```

## Stage 1: score the trait (brms) and the distal (CFA)

The trait is the random intercept of a Gaussian brms model with a **random
sigma** (a subject-specific residual standard deviation):

```{r stage1, eval = FALSE}
# The nested form (1 | p | Subject) lets brms estimate the *joint* (correlated)
# covariance of the location and scale random effects, which get_fs() reads.
fit <- brm(bf(y ~ 1 + (1 | p | Subject), sigma ~ (1 | p | Subject)),
           data = d, family = gaussian(),
           chains = 2, iter = 3000, seed = 1, silent = TRUE)
fs <- get_fs(fit)

# The distal construct is scored by a three-indicator CFA.
fs_d <- get_fs(dj, model = "distal =~ d1 + d2 + d3")
```

The brms model has two coefficients — the location (the trait, `u0`) and the
scale (the noise level, `u1`) — so it contributes two scores, `fs_u0` and
`fs_u1`, each with its own standard error; the CFA contributes the distal score
`fs_distal`. The MCMC fit and the `get_fs()` calls are expensive, so the results
are cached; the analysis below loads them:

```{r load, echo = FALSE, include = FALSE}
x <- readRDS("sim_brms_ls_vignette.rds")
fs <- x$fs; fs_d <- x$fs_d; beta1 <- x$beta1; re_summary <- x$re_summary
```

The random sigma shows up directly in the stage-1 output: the standard error of
the location score `fs_u0` **differs from subject to subject** —

```{r scores}
head(fs[, c("fs_u0", "fs_u1", "fs_u0_se", "fs_u1_se")])
range(fs$fs_u0_se)
```

Subjects whose residual variance is large have a wide `fs_u0_se` (their score is
less certain); those with a small residual variance have a narrow one. A brms
model *without* a random sigma would return a single, constant standard error
for every subject. The fitted random-effects hyperparameters confirm that both
the location and the scale random effects vary by subject — in particular,
`sd_Subject__sigma_Intercept` is the standard deviation of the (log) subject-
specific residual standard deviation:

```{r re}
re_summary
```

## Stage 2a: the standard route — `tspa()` (pooled error)

We relate the distal construct to the latent trait (`u0`, the location score).
Both are scored latents, so the naive regression plugs the two *noisy scores*
into a regression, ignoring their measurement error:

```{r stage2}
df <- data.frame(fs_u0 = fs$fs_u0, fs_u0_se = fs$fs_u0_se,
                 fs_distal = fs_d$fs_distal, fs_distal_se = fs_d$fs_distal_se)

## (a) Naive: plug the scores into a regression, ignoring their measurement error
naive_fit <- lm(fs_distal ~ fs_u0, data = df)
coef(naive_fit)

## (b) 2S-PA (standard): correct for the measurement error
tspa_fit <- tspa("distal ~ u0", data = df)
coef(tspa_fit)
```

`R2spa`'s standard stage 2 is a `lavaan` model, which represents each
construct's measurement error with a **single** variance. Given person-specific
standard errors, it pools them into one value:

```{r model}
cat(attributes(tspa_fit)$tspaModel)
```

## Stage 2b: the exact route — `tspa_mx_model()` (per-subject error)

`lavaan` cannot hold a different error variance per observation, so the standard
`tspa()` pools the person-specific standard errors. `tspa_mx_model()` (an
`OpenMx` model) can: it takes each subject's measurement-error variance as a
**definition variable**, a per-observation quantity that varies by row. This is
the exact, non-pooled correction (Lai & Hsiao, 2022):

```{r mx, warning = FALSE}
library(OpenMx)
comb <- cbind(fs, fs_d)
# unit loadings; the error variances are per-subject definition variables
# (columns of `comb` that tspa_mx_model() reads row by row):
fsL <- matrix(c(1, NA, NA, 1), nrow = 2,
              dimnames = list(c("fs_u0", "fs_distal"), c("u0", "distal")))
fsT <- matrix(c("ev_fs_u0", NA, NA, "ev_fs_distal"), nrow = 2,
              dimnames = list(c("fs_u0", "fs_distal"), c("fs_u0", "fs_distal")))
mx_fit <- tspa_mx_model("distal ~ u0", data = comb, fsL = fsL, fsT = fsT)
# read the structural path (OpenMx reports it in the A matrix)
pp <- summary(mx_fit)$parameters
p2 <- pp[pp$matrix == "A" & pp$row == "distal" & pp$col == "u0",
         c("row", "col", "Estimate", "Std.Error")]
print(p2)
```

The `fsT` above is a character matrix: each cell names a column of `comb` that
carries that subject's error variance. `ev_fs_u0` **varies from subject to
subject** (the random-sigma signature); `ev_fs_distal` is constant (the CFA is
homoskedastic). A `lavaan` model would fix `fs_u0`'s error to a single pooled
value; the `OpenMx` model applies each subject's value.

## Comparison

```{r compare}
pt <- lavaan::parameterTable(tspa_fit)
i  <- pt$lhs == "distal" & pt$op == "~" & pt$rhs == "u0"
est <- c(true    = beta1,
         naive   = unname(coef(naive_fit)["fs_u0"]),
         "tspa"  = unname(pt$est[i]),
         "tspa_mx" = unname(p2$Estimate[1]))
se  <- c(true    = NA,
         naive   = summary(naive_fit)$coefficients["fs_u0", "Std. Error"],
         "tspa"  = unname(pt$se[i]),
         "tspa_mx" = unname(p2$Std.Error[1]))
print(round(data.frame(estimate = est, SE = se), 3), row.names = TRUE)
```

The naive regression is **attenuated toward zero** (0.928 versus the true 1.0):
because the score `fs_u0` is a noisy estimate of the trait, regressing on it
dilutes the slope (errors-in-variables bias; Cole & Preacher, 2014). Both 2S-PA
routes correct for the measurement error and recover the true coefficient
(`tspa` 1.039, `tspa_mx_model` 1.020). Here the difference between the pooled
(`tspa`) and per-subject (`tspa_mx_model`) corrections is small — each subject
has several measurements, so the pooled average is a good stand-in for the
person-specific value — but the per-subject route is the exact one and is the
right choice as the error becomes more heteroskedastic (fewer measurements per
subject, or a wider spread of residual variances).

## Why the random sigma matters

A random sigma is what makes the measurement error *person-specific*: `get_fs()`
returns a different `fs_u0_se` for each subject. The standard `tspa()` (a
`lavaan` model) represents the error with a single variance per construct, so it
pools these person-specific standard errors into one value — a reasonable
approximation when the scores come from a common measurement process, but a
simplification when the error is strongly heteroskedastic. `tspa_mx_model()`
(stage 2b above) is the exact, per-subject alternative: it supplies each
subject's error variance as a definition variable, so no pooling is needed.

## Scope and notes

- `brms` is a `Suggests` dependency; `get_fs()` stops with a clear message if it
  is not installed.
- For brms location-scale models only the **EB** (posterior) scoring is
  available; `method = "ML"` is rejected, because the prior-free OLS scoring
  assumes a constant residual variance, which a random sigma violates.
- `get_fs()` supports a **single** random-effects grouping factor; multiple
  distinct grouping factors are not yet supported.
- The nested `(1 | p | Subject)` form is what makes brms estimate the joint
  location–scale covariance. A model with independent location and scale random
  effects (no shared random-effects term) is not currently supported.
- `tspa_mx_model()` (Phase 1) restricts every variable in the structural model
  to a corrected latent (one with a factor score), which is why the distal
  outcome here is scored as a latent rather than entered as a single observed
  variable.

## References

- Cole, D. A., & Preacher, K. J. (2014). Manifest variable path analysis:
  Potentially serious and misleading consequences due to uncorrected measurement
  error. *Psychological Methods, 19*(2), 300–315.
  [https://doi.org/10.1037/a0033805](https://doi.org/10.1037/a0033805)
- Bürkner, P.-C. (2017). brms: An R package for Bayesian multilevel models using
  Stan. *Journal of Statistical Software, 80*(1), 1–28.
  [https://doi.org/10.18637/jss.v080.i01](https://doi.org/10.18637/jss.v080.i01)
- Lai, M. H. C., & Hsiao, Y.-Y. (2022). Two-stage path analysis with definition
  variables: An alternative framework to account for measurement error.
  *Psychological Methods, 27*(4), 568–588.
  [https://doi.org/10.1037/met0000410](https://doi.org/10.1037/met0000410)
- Lai, M. H. C., & Liu, S. (2026). A two-stage approach to account for
  measurement error when using empirical Bayes estimates of random slopes.
  *Psychological Methods.* Advance online publication.
  [https://doi.org/10.1037/met0000838](https://doi.org/10.1037/met0000838)
- Neale, M. C., et al. (2016). OpenMx: An open source extended structural
  equation modeling framework. *Psychometrika, 81*(2), 535–549.
  [https://doi.org/10.1007/s11336-016-9549-x](https://doi.org/10.1007/s11336-016-9549-x)
- Rosseel, Y. (2012). *lavaan*: An R package for structural equation modeling.
  *Journal of Statistical Software, 48*(2), 1–36.
  [https://doi.org/10.18637/jss.v048.i02](https://doi.org/10.18637/jss.v048.i02)
