---
title: "Multi-Factor Measurement Model (OpenMx)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tspa-vignette-mx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message = FALSE}
library(lavaan)
library(R2spa)
library(OpenMx)
library(mirt)
```

Two-stage path analysis (2S-PA) can be fit with `OpenMx` as well as the
default `lavaan` backend. `tspa_mx_model()` takes a `lavaan`-style structural
model string plus the (per-row) factor-score loading / error information from
`get_fs()` and returns a fitted `OpenMx` RAM model.

# With a Linear Factor Model

The example is from https://lavaan.ugent.be/tutorial/sem.html.

```{r}
# CFA
my_cfa <- "
  # latent variables
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
"
(fs_dat <- get_fs(PoliticalDemocracy, model = my_cfa, std.lv = TRUE,
                  format = "list")) |> head()
```

```{r}
tspa_fit <- tspa(model = "dem60 ~ ind60", data = fs_dat,
                 fsT = attr(fs_dat, "fsT"),
                 fsL = attr(fs_dat, "fsL"))
parameterEstimates(tspa_fit)
```

## Using OpenMx

The measurement information is read straight from the `fsL` / `fsT`
attributes of the `get_fs()` output. The score columns (`fs_ind60`,
`fs_dem60`) and the latent names (`ind60`, `dem60`) must already be present
in `data` / the model string.

```{r}
tspa_mx <- tspa_mx_model(model = "dem60 ~ ind60", data = fs_dat,
                         fsL = attr(fs_dat, "fsL"),
                         fsT = attr(fs_dat, "fsT"))
# Run OpenMx
tspa_mx_fit <- mxRun(tspa_mx)
# Summarize the results
summary(tspa_mx_fit)
```

### With Mean Structure

```{r}
my_cfa <- "
  # latent variables
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
    x1 + y1 ~ 0
    ind60 + dem60 ~ 1
"
fs_dat <- get_fs(PoliticalDemocracy, model = my_cfa,
                 meanstructure = TRUE,
                 format = "list")
```

lavaan

```{r}
tspa_fit <- tspa(model = "dem60 ~ ind60\ndem60 + ind60 ~ 1", data = fs_dat,
                 fsT = attr(fs_dat, "fsT"),
                 fsL = attr(fs_dat, "fsL"),
                 fsb = attr(fs_dat, "fsb"))
parameterEstimates(tspa_fit, standardized = TRUE)
```

OpenMx

```{r}
tspa_mx <- tspa_mx_model(model = "dem60 ~ ind60\ndem60 + ind60 ~ 1",
                         data = fs_dat,
                         fsL = attr(fs_dat, "fsL"),
                         fsT = attr(fs_dat, "fsT"),
                         fsb = attr(fs_dat, "fsb"))
# Run OpenMx
tspa_mx_fit <- mxRun(tspa_mx)
# Summarize the results
summary(tspa_mx_fit)
```

## Compare to joint model

```{r}
jreg <- sem(
    "
    # latent variables
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
    # latent regression
    dem60 ~ ind60
    ",
    data = PoliticalDemocracy, std.lv = TRUE)
coef(jreg)["dem60~ind60"]
```

# Combined with IRT

Example from Lai & Hsiao (2022, Psychological Methods)

## Not accounting for error

```{r}
# Simulate data with mirt
set.seed(1235)
num_obs <- 1000
# Simulate theta
eta <- MASS::mvrnorm(num_obs, mu = c(0, 0), Sigma = diag(c(1, 1 - 0.5^2)),
                     empirical = TRUE)
th1 <- eta[, 1]
th2 <- -1 + 0.5 * th1 + eta[, 2]
# items and response data
a1 <- matrix(1, 10)
d1 <- matrix(rnorm(10))
a2 <- matrix(runif(10, min = 0.5, max = 1.5))
d2 <- matrix(rnorm(10))
dat1 <- simdata(a = a1, d = d1, N = num_obs, itemtype = "2PL", Theta = th1)
dat2 <- simdata(a = a2, d = d2, N = num_obs, itemtype = "2PL", Theta = th2)
# Factor scores
mod1 <- mirt(dat1, model = 1, itemtype = "Rasch", verbose = FALSE)
mod2 <- mirt(dat2, model = 1, itemtype = "2PL", verbose = FALSE)
fs1 <- fscores(mod1, full.scores.SE = TRUE)
fs2 <- fscores(mod2, full.scores.SE = TRUE)
lm(fs2[, 1] ~ fs1[, 1])  # attenuated coefficient
```

## Factor-score intercepts via get_fs()

The `mirt` fit above is also scored directly by `get_fs()`, which returns the EAP
scores together with their per-row standard errors, implied loadings and error
variances (and, via `fs_indiv()`, the definition variables) — so the hand-rolled
`fs_eap` frame in the next section is optional. A factor prior mean can be given
with `prior_mean` (mirt's default is a zero-mean factor). When the mean is not
zero, the per-row factor-score intercepts stored in the `fsb` attribute are
non-zero: each `fsb_i` is the prior mean scaled by that row's shrinkage factor
(the posterior variance `Vpost_i`), so the intercepts — and the score means —
track the prior mean, and vary from row to row.

```{r}
fs  <- get_fs(mod1)
fse <- get_fs(mod1, prior_mean = c(F1 = 2))
head(fs, 3)                                   # score / SE / loading / error variance
round(c(fsb.r1.default = unname(attr(fs, "fsb")[[1]]),
        fsb.r2.default = unname(attr(fs, "fsb")[[2]]),
        fsb.r1.prior2  = unname(attr(fse, "fsb")[[1]]),
        fsb.r2.prior2  = unname(attr(fse, "fsb")[[2]])), 4)
round(c(score.mean.default = mean(fs$fs_F1),
        score.mean.prior2  = mean(fse$fs_F1)), 4)
```

Under mirt's default zero-mean prior every `fsb` is zero (the scores are mean
zero); with `prior_mean = c(F1 = 2)` they are non-zero and differ by row. These
`fsb` values are exactly what `tspa_mx_model()` reads as its score-intercept row
("With Mean Structure" above).

## 2S-PA with EAP (shrinkage) scores

Because EAP (shrinkage) scores are used, we use the following properties for
these scores ($\tilde \eta_i$ for person $i$):

- Standard error (SE, from software output) = $\sqrt{V(\eta) (1 - \rho_{\tilde \eta, i})}$, where $\rho_{\tilde \eta, i}$ is the reliability of $\tilde \eta_i$
- Loading of $\tilde \eta_i$ on $\eta$ is equal to $\rho_{\tilde \eta, i}$ = $1 - \text{SE}^2 / V(\eta)$
- Standard error of measurement for $\tilde \eta_i$ is equal to $\rho_{\tilde \eta, i} \text{SE}$
- Based on the above, the total variance of $\tilde \eta_i$ is also $\rho_{\tilde \eta, i}$.

```{r}
# Combine into data set
fs_eap <- cbind(fs1, fs2) |>
    as.data.frame() |>
    setNames(c("fs1", "se_fs1", "fs2", "se_fs2")) |>
    # Compute reliability and error variances
    within(expr = {
        rel_fs1 <- 1 - se_fs1^2
        rel_fs2 <- 1 - se_fs2^2
        ev_fs1 <- se_fs1^2 * (1 - se_fs1^2)
        ev_fs2 <- se_fs2^2 * (1 - se_fs2^2)
    })
```

The per-person reliability and error variance are supplied as *definition
variables*: each `fsL` / `fsT` cell holds the name of the `data` column that
carries that person's value. The structural latent names (`f1`, `f2`) are kept
distinct from the score columns (`fs1`, `fs2`).

```{r}
cross_load <- matrix(c("rel_fs1", NA, NA, "rel_fs2"), nrow = 2) |>
    `dimnames<-`(list(c("fs1", "fs2"), c("f1", "f2")))
err_cov <- matrix(c("ev_fs1", NA, NA, "ev_fs2"), nrow = 2) |>
    `dimnames<-`(rep(list(c("fs1", "fs2")), 2))
tspa_mx <- tspa_mx_model(model = "f2 ~ f1\nf2 + f1 ~ 1",
                         data = fs_eap, fsL = cross_load, fsT = err_cov)
# Run OpenMx
tspa_mx_fit <- mxRun(tspa_mx)
# Summarize the results
summary(tspa_mx_fit)
```

## Joint model reference

The gold-standard comparison is a joint (one-step) MIMIC/SEM fit. With 2PL
items this uses the polychoric-WLS approximation here; a full FIML fit is
computationally prohibitive at this scale.

```{r}
dat <- cbind(dat1, dat2)
colnames(dat) <- paste0("i", 1:20)
wls_fit <- sem("
f1 =~ i1 + i2 + i3 + i4 + i5 + i6 + i7 + i8 + i9 + i10
f2 =~ i11 + i12 + i13 + i14 + i15 + i16 + i17 + i18 + i19 + i20
f2 ~ f1
", data = dat, ordered = TRUE, std.lv = TRUE)
coef(wls_fit)["f2~f1"]
```

## Multidimensional Measurement Model

When a multidimensional model is used, for EAP scores we use the following
properties generalized from the unidimensional case. Specifically, let
$V(\boldsymbol \eta)$ = $\boldsymbol{\psi}$.

- Loading of $\tilde{\boldsymbol \eta}_i$ on $\boldsymbol \eta$ is $\boldsymbol{\Lambda}_{\tilde \eta, i}$, which is also the reliability matrix (i.e., the squared covariance between $\tilde{\boldsymbol \eta}_i$ and $\boldsymbol \eta$).
- Error matrix ($\mathbf{E}$, from software output) = $(\mathbf{I} - \boldsymbol{\Lambda}_{\tilde \eta, i})\boldsymbol{\psi}$. Therefore, $\boldsymbol{\Lambda}_{\tilde \eta, i} = \mathbf{I} - \mathbf{E} \boldsymbol{\psi}^{-1}$.
- Error variance of $\tilde \eta_i$ is $\boldsymbol{\Lambda}_{\tilde \eta, i} \mathbf{E}$.

```{r}
dat <- cbind(dat1, dat2)
colnames(dat) <- paste0("Item_", 1:20)
# Multidimensional IRT
mod <- "
F1 = 1-10
F2 = 11-20
COV = F1*F2
CONSTRAIN = (1-10, a1)
"
mfit <- mirt(dat, model = mod, itemtype = "2PL", verbose = FALSE)
```

```{r}
# Factor scores, plus the per-row loadings and error covariances get_fs()
# attaches (the same full-factor-covariance correction the hand-rolled form
# below used to compute by hand).
fs_dat <- get_fs(mfit)
```

```{r}
# Structural model on the two latent dimensions (distinct from the score columns)
S <- c("fs_F1", "fs_F2")
cross_load <- matrix(c("F1_by_fs_F1", "F1_by_fs_F2",
                       "F2_by_fs_F1", "F2_by_fs_F2"), nrow = 2) |>
    `dimnames<-`(list(S, c("theta1", "theta2")))
err_cov <- matrix(c("ev_fs_F1", "ecov_fs_F2_fs_F1",
                    "ecov_fs_F2_fs_F1", "ev_fs_F2"), nrow = 2) |>
    `dimnames<-`(rep(list(S), 2))
tspa_mx2 <- tspa_mx_model(model = "theta2 ~ theta1\ntheta2 + theta1 ~ 1",
                          data = fs_dat, fsL = cross_load, fsT = err_cov)
# Run OpenMx
tspa_mx2_fit <- mxRun(tspa_mx2)
# Summarize the results
summary(tspa_mx2_fit)
```
