---
title: "2S-PA with a brms location-scale model (random sigma)"
author: "Mark Lai"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{2S-PA with a brms location-scale model (random sigma)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Two-stage path analysis (2S-PA; Lai & Hsiao, 2022) estimates a structural model
whose constructs are measured with error in two steps. **Stage 1** scores each
construct and records its measurement error; **stage 2** path-analyzes the
scores while correcting for that error, so the structural coefficients are not
biased by it (Cole & Preacher, 2014). In `R2spa`, `get_fs()` is stage 1 and
`tspa()` is stage 2.

This vignette uses stage 1 from a **brms location-scale mixed model** — a
Gaussian model in which the residual (measurement) scale has its own random
effect, a *random sigma*. A random sigma makes the measurement error
**person-specific** (heteroskedastic): subjects with a larger residual variance
have noisier scores. `get_fs()` recovers each subject's score *and* its
person-specific standard error.

To make the comparison concrete, we simulate data with **known** structural
coefficients — the distal construct is driven by both the latent trait and the
noise level (the random sigma) — fit the brms location-scale model (and a small
CFA for the distal construct), and regress the distal construct on the trait
*and* the noise level in three ways at stage 2: a naive regression that ignores
the measurement error, the standard `tspa()` (a `lavaan` model that pools the
person-specific error into one value), and `tspa_mx_model()` (an `OpenMx` model
that uses the error **per subject**, via definition variables — the exact,
non-pooled route).

## The data

We simulate `n` subjects, each nested within one of `n_sites` sites, with `m`
noisy measurements per subject. Each subject has a latent trait (the random
intercept) and a person-specific noise level (the random sigma); the measured
values are the trait plus person-specific noise. A **distal construct** (one per
subject) is generated from *both* the subject's true trait (coefficient `beta1`)
and its noise level (coefficient `beta2`, on the log-sigma scale that the sigma
score recovers), and is measured by three noisy indicators, so it is itself
scored as a latent.

```{r dgp, eval = FALSE}
suppressMessages({ library(brms); library(R2spa) })
set.seed(33)

n_sites <- 20; per_site <- 10; n <- n_sites * per_site; m <- 5
beta1 <- 1.0   # effect of the trait on the distal
beta2 <- 1.0   # effect of the noise level (the random sigma) on the distal
p      <- factor(rep(1:n_sites, each = per_site))  # site (higher level)
b0     <- rnorm(n, 0, 0.5)                         # true latent trait
sig    <- exp(rnorm(n, -1.0, 0.5))                 # person-specific noise SD
sid    <- factor(rep(1:n, each = m))
y      <- b0[as.integer(sid)] + sig[as.integer(sid)] * rnorm(n * m)
d      <- data.frame(y = y, Subject = sid, p = p[as.integer(sid)])
# distal construct, driven by BOTH the trait and the noise level, measured by
# three noisy indicators (log(sig) is the scale RE that the sigma score recovers)
d0     <- beta1 * b0 + beta2 * log(sig) + rnorm(n, 0, 0.5)
dj     <- data.frame(d1 = d0 + rnorm(n, 0, 0.5),
                     d2 = d0 + rnorm(n, 0, 0.5),
                     d3 = d0 + rnorm(n, 0, 0.5))
```

## Stage 1: score the trait (brms) and the distal (CFA)

The trait is the random intercept of a Gaussian brms model with a **random
sigma** (a subject-specific residual standard deviation):

```{r stage1, eval = FALSE}
# The nested form (1 | p | Subject) lets brms estimate the *joint* (correlated)
# covariance of the location and scale random effects, which get_fs() reads.
fit <- brm(bf(y ~ 1 + (1 | p | Subject), sigma ~ (1 | p | Subject)),
           data = d, family = gaussian(),
           chains = 2, iter = 3000, seed = 1, silent = TRUE)
fs <- get_fs(fit)

# The distal construct is scored by a three-indicator CFA.
fs_d <- get_fs(dj, model = "distal =~ d1 + d2 + d3")
```

The brms model has two coefficients — the location (the trait, `u0`) and the
scale (the noise level, `u1`) — so it contributes two scores, `fs_u0` and
`fs_u1`, each with its own standard error; the CFA contributes the distal score
`fs_distal`. The MCMC fit and the `get_fs()` calls are expensive, so the results
are cached; the analysis below loads them:

```{r load, echo = FALSE, include = FALSE}
library(lavaan)
library(R2spa)
x <- readRDS("sim_brms_ls_vignette.rds")
fs <- x$fs; fs_d <- x$fs_d; beta1 <- x$beta1; beta2 <- x$beta2
re_summary <- x$re_summary
```

The random sigma shows up directly in the stage-1 output: the standard error of
the location score `fs_u0` **differs from subject to subject** —

```{r scores}
head(fs[, c("fs_u0", "fs_u1", "fs_u0_se", "fs_u1_se")])
range(fs$fs_u0_se)
```

Subjects whose residual variance is large have a wide `fs_u0_se` (their score is
less certain); those with a small residual variance have a narrow one. A brms
model *without* a random sigma would return a single, constant standard error
for every subject. The fitted random-effects hyperparameters confirm that both
the location and the scale random effects vary by subject — in particular,
`sd_Subject__sigma_Intercept` is the standard deviation of the (log) subject-
specific residual standard deviation:

```{r re}
re_summary
```

## Stage 2a: the standard route — `tspa()` (pooled error)

We now relate the distal construct to **both** the trait (`u0`, the location
score) and the noise level (`u1`, the random-sigma score). All three are scored
latents, so the naive regression plugs the three *noisy scores* into a
regression, ignoring their measurement error.

The two brms scores come from the same location-scale model, so their
measurement errors are **correlated** (the location and scale random effects are
estimated jointly); the distal score, from a separate CFA, has no error
covariance with them. `tspa()` expresses this with a **block-diagonal** error
matrix `fsT` — the two predictor scores share one error (co)variance block, the
distal its own — and a **non-diagonal** loading matrix `fsL` (the two scores
cross-load on each other's latent, so the loadings are not unit and not
diagonal). The per-subject quantities are pooled to one representative matrix
(the mean):

```{r stage2}
L_b <- apply(attr(fs, "fsL"), c(1, 2), mean)   # 2 x 2, the brms loading block
T_b <- apply(attr(fs, "fsT"), c(1, 2), mean)   # 2 x 2, error block incl. the error covariance
L_d <- attr(fs_d, "fsL")[[1]]                  # 1 x 1, the CFA loading
T_d <- attr(fs_d, "fsT")[[1]]                  # 1 x 1, the CFA error variance
scores  <- c("fs_u0", "fs_u1", "fs_distal")
latents <- c("u0", "u1", "distal")
# block-diagonal: the predictor block shares its loadings and its error
# (co)variance; the distal is its own block (zero error covariance with the
# predictors)
fsL <- matrix(0, 3, 3, dimnames = list(scores, latents))
fsL[1:2, 1:2] <- L_b; fsL[3, 3] <- L_d
fsT <- matrix(0, 3, 3, dimnames = list(scores, scores))
fsT[1:2, 1:2] <- T_b; fsT[3, 3] <- T_d
comb <- cbind(fs, fs_d)

## (a) Naive: plug the scores into a regression, ignoring their measurement error
naive_fit <- lm(fs_distal ~ fs_u0 + fs_u1, data = comb)
coef(naive_fit)

## (b) 2S-PA (standard): correct for the (correlated) measurement error
tspa_fit <- tspa("distal ~ u0 + u1", data = comb, fsT = fsT, fsL = fsL)
coef(tspa_fit)
```

The rendered model fixes the (pooled) loadings, the per-block error variances,
and — the key point once the sigma is a predictor — the error covariance between
the two predictor scores (`fs_u1 ~~ <cov> * fs_u0`), while the (true) latent
variances and covariances stay free:

```{r model}
cat(attributes(tspa_fit)$tspaModel)
```

## Stage 2b: the exact route — `tspa_mx_model()` (per-subject error)

`lavaan` cannot hold a different error (co)variance per observation, so the
standard `tspa()` pools the person-specific error to one value. `tspa_mx_model()`
(an `OpenMx` model) can: it takes each subject's measurement-error variance
**and** the error covariance between the two predictor scores as **definition
variables**, per-observation quantities that vary by row. This is the exact,
non-pooled correction (Lai & Hsiao, 2022):

```{r mx, warning = FALSE}
library(OpenMx)
# the loadings are fixed at the pooled values (above); each subject's error
# (co)variances are per-row definition variables — the predictor block's error
# covariance (ecov_fs_u1_fs_u0) included, the distal's block separate
fsT <- matrix(NA_character_, 3, 3, dimnames = list(scores, scores))
fsT[1, 1] <- "ev_fs_u0"
fsT[2, 1] <- "ecov_fs_u1_fs_u0"
fsT[2, 2] <- "ev_fs_u1"
fsT[3, 3] <- "ev_fs_distal"
mx_fit <- tspa_mx_model("distal ~ u0 + u1", data = comb, fsL = fsL, fsT = fsT)
# read the structural paths (OpenMx reports them in the A matrix)
pp <- summary(mx_fit)$parameters
p2 <- pp[pp$matrix == "A" & pp$row == "distal" & pp$col %in% c("u0", "u1"),
         c("row", "col", "Estimate", "Std.Error")]
print(p2)
```

The `fsT` above is a character matrix: each cell names a column of `comb` that
carries that subject's error quantity. `ev_fs_u0`, `ev_fs_u1`, and
`ecov_fs_u1_fs_u0` **vary from subject to subject** (the random-sigma
signature); `ev_fs_distal` is constant (the CFA is homoskedastic). A `lavaan`
model would fix the predictor error block to a single pooled value; the `OpenMx`
model applies each subject's value.

## Comparison

```{r compare}
pt <- lavaan::parameterTable(tspa_fit)
i0 <- pt$lhs == "distal" & pt$op == "~" & pt$rhs == "u0"
i1 <- pt$lhs == "distal" & pt$op == "~" & pt$rhs == "u1"
out <- data.frame(
  method  = c("true", "naive", "tspa", "tspa_mx"),
  est_u0  = c(beta1, coef(naive_fit)["fs_u0"], pt$est[i0], p2$Estimate[p2$col == "u0"]),
  se_u0   = c(NA, summary(naive_fit)$coefficients["fs_u0", "Std. Error"],
              pt$se[i0], p2$Std.Error[p2$col == "u0"]),
  est_u1  = c(beta2, coef(naive_fit)["fs_u1"], pt$est[i1], p2$Estimate[p2$col == "u1"]),
  se_u1   = c(NA, summary(naive_fit)$coefficients["fs_u1", "Std. Error"],
              pt$se[i1], p2$Std.Error[p2$col == "u1"]))
out[, -1] <- round(out[, -1], 3)
print(out, row.names = FALSE)
```

Both predictors have a true coefficient of 1. The naive regression, which treats
the noisy scores as error-free, is biased — here **inflated** (1.09 and 1.22),
because the two predictors' measurement errors are correlated, so part of the
shared error is misattributed to the slopes. The standard `tspa()` route, which
pools the person-specific error to a single average block, is biased here too,
and more so for the sigma (1.29): with a strongly person-specific error, one
average block is a poor stand-in for the subject-specific one. The per-subject
`tspa_mx_model()` route, which applies each subject's error variances *and* the
error covariance, recovers both true coefficients (1.01 and 1.05).

This is the regime where the per-subject route is essential: the random sigma
makes the measurement error strongly person-specific, and once two correlated
scores share that error, the pooled correction (a single average block) no
longer approximates it well. With a single predictor the pooled route is a good
approximation (as in the single-factor case), but the correlated two-predictor
case stresses it.

## Why the random sigma matters

A random sigma is what makes the measurement error *person-specific*: `get_fs()`
returns a different `fs_u0_se` (and `fs_u1_se`) for each subject, and — because
the location and scale random effects are estimated jointly — a person-specific
error *covariance* between the two scores (`ecov_fs_u1_fs_u0`). The standard
`tspa()` (a `lavaan` model) represents the error with a single (pooled) variance
per construct and a single (pooled) covariance, so it pools these person-specific
quantities into one block — a reasonable approximation when the error is only
mildly heteroskedastic, but a poor one when the random sigma makes it strongly
person-specific, as here. `tspa_mx_model()` (stage 2b above) is the exact,
per-subject alternative: it supplies each subject's error variances and the
error covariance as definition variables, so no pooling is needed.

## Scope and notes

- `brms` is a `Suggests` dependency; `get_fs()` stops with a clear message if it
  is not installed.
- For brms location-scale models only the **EB** (posterior) scoring is
  available; `method = "ML"` is rejected, because the prior-free OLS scoring
  assumes a constant residual variance, which a random sigma violates.
- `get_fs()` supports a **single** random-effects grouping factor; multiple
  distinct grouping factors are not yet supported.
- The nested `(1 | p | Subject)` form is what makes brms estimate the joint
  location–scale covariance. A model with independent location and scale random
  effects (no shared random-effects term) is not currently supported.
- `tspa_mx_model()` (Phase 1) restricts every variable in the structural model
  to a corrected latent (one with a factor score), which is why the distal
  outcome here is scored as a latent rather than entered as a single observed
  variable.
- The standard `tspa()` route corrects for the error **covariance** between the
  two predictor scores through the off-diagonal of the (block-diagonal) `fsT`
  matrix; the single-factor `se_fs` form of `tspa()` expresses only a single
  variance per construct and cannot represent that covariance.

## References

- Cole, D. A., & Preacher, K. J. (2014). Manifest variable path analysis:
  Potentially serious and misleading consequences due to uncorrected measurement
  error. *Psychological Methods, 19*(2), 300–315.
  [https://doi.org/10.1037/a0033805](https://doi.org/10.1037/a0033805)
- Bürkner, P.-C. (2017). brms: An R package for Bayesian multilevel models using
  Stan. *Journal of Statistical Software, 80*(1), 1–28.
  [https://doi.org/10.18637/jss.v080.i01](https://doi.org/10.18637/jss.v080.i01)
- Lai, M. H. C., & Hsiao, Y.-Y. (2022). Two-stage path analysis with definition
  variables: An alternative framework to account for measurement error.
  *Psychological Methods, 27*(4), 568–588.
  [https://doi.org/10.1037/met0000410](https://doi.org/10.1037/met0000410)
- Lai, M. H. C., & Liu, S. (2026). A two-stage approach to account for
  measurement error when using empirical Bayes estimates of random slopes.
  *Psychological Methods.* Advance online publication.
  [https://doi.org/10.1037/met0000838](https://doi.org/10.1037/met0000838)
- Neale, M. C., et al. (2016). OpenMx: An open source extended structural
  equation modeling framework. *Psychometrika, 81*(2), 535–549.
  [https://doi.org/10.1007/s11336-016-9549-x](https://doi.org/10.1007/s11336-016-9549-x)
- Rosseel, Y. (2012). *lavaan*: An R package for structural equation modeling.
  *Journal of Statistical Software, 48*(2), 1–36.
  [https://doi.org/10.18637/jss.v048.i02](https://doi.org/10.18637/jss.v048.i02)
