---
title: "Diagnostic plots for `tspa` models"
author: "Jimmy Zhang"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{tspa-diagnostic-plot-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The example is from <https://lavaan.ugent.be/tutorial/sem.html>.

Load packages

```{r load-pkg, message = FALSE}
library(lavaan)
library(R2spa)
```

## Single group, single predictor

```{r}
model <- ' 
  # latent variable definitions
     ind60 =~ x1 + x2 + x3
     dem60 =~ y1 + a*y2 + b*y3 + c*y4

  # regressions
    dem60 ~ ind60
'

fs_dat_ind60 <- get_fs(data = PoliticalDemocracy, 
                       model = "ind60 =~ x1 + x2 + x3")
fs_dat_dem60 <- get_fs(data = PoliticalDemocracy, 
                       model = "dem60 =~ y1 + y2 + y3 + y4")
fs_dat <- cbind(fs_dat_ind60, fs_dat_dem60)

tspa_fit_1 <- tspa(model = "dem60 ~ ind60", 
                   data = fs_dat, 
                   se_fs = list(ind60 = 0.1213615, dem60 = 0.6756472),
                   meanstructure = T)
```

We recommend researchers to examine the diagnostic plot of factor scores in the model. They can use the function `tspa_plot()` to obtain two plots: (a) a scatter plot between factors, (b) a residual plot between factors.

Features of `tspa_plot()`:

-   The function is able to pass arguments to the base R `plot()` function.
-   Users can define the title of the scatter plot and label names of axis.
-   Abbreviation argument allows users to choose whether using abbreviated group names.
-   Users can choose whether generating the plot one by one by hitting the <return> on the keyboard.

```{r example_single_group_single_factor}
# par(mar = c(2,2,3,2))
tspa_plot(tspa_fit_1,
          col = "blue",
          cex.lab = 1.2,
          cex.axis = 1,
          fscores_type = "original",
          ask = TRUE)
```

### Using estimates from `tspa()` function

```{r include=FALSE}
latent_estimates <- data.frame(lavPredict(tspa_fit_1))
plot(latent_estimates$ind60, 
     latent_estimates$dem60, 
     main = "Scatterplot", 
     xlab = "ind60", 
     ylab = "dem60", 
     pch = 16, 
     ps = 1.25, 
     cex.lab = 1.2, 
     cex.axis = 1.2)
slope <- coef(tspa_fit_1)["dem60~ind60"]
abline(a = unname(coef(tspa_fit_1)["fs_ind60~1"] - slope*coef(tspa_fit_1)["fs_dem60~1"]), 
       b = slope, 
       lwd = 2)
```

## Single group, multiple predictors

```{r}
model <- ' 
  # latent variable definitions
     ind60 =~ x1 + x2 + x3
     dem60 =~ y1 + y2 + y3 + y4
     dem65 =~ y5 + y6 + y7 + y8

  # regressions
    dem60 ~ ind60
    dem65 ~ ind60 + dem60

  # # residual correlations
  #   y1 ~~ y5
  #   y2 ~~ y4 + y6
  #   y3 ~~ y7
  #   y4 ~~ y8
  #   y6 ~~ y8
'

fs_dat_ind60 <- get_fs(data = PoliticalDemocracy, 
                       model = "ind60 =~ x1 + x2 + x3")
fs_dat_dem60 <- get_fs(data = PoliticalDemocracy, 
                       model = "dem60 =~ y1 + y2 + y3 + y4")
fs_dat_dem65 <- get_fs(data = PoliticalDemocracy, 
                       model = "dem65 =~ y5 + y6 + y7 + y8")
fs_dat <- cbind(fs_dat_ind60, fs_dat_dem60, fs_dat_dem65)

tspa_fit_2 <- tspa(model = "dem60 ~ ind60
                            dem65 ~ ind60 + dem60", 
                   data = fs_dat, 
                   se_fs = list(ind60 = 0.1213615, dem60 = 0.6756472, 
                                dem65 = 0.5724405))
```

```{r example_single_group_multiple_factors}
# Title, xlab, and ylab each with same names.
tspa_plot(tspa_fit_2,
          ps = 10,
          col = "blue",
          cex.lab = 1.2,
          cex.axis = 1)

# Title, xlab, and ylab each with separate names.
tspa_plot(tspa_fit_2,
          ps = 10,
          col = "darkgray",
          cex.lab = 1.2,
          cex.axis = 1,
          title = c("Scatterplot_I", "Scatterplot_II", "Scatterplot_III"),
          label_x = c("factor_1", "factor_2", "factor_3"),
          label_y = c("factor_a", "factor_b", "factor_c"))
```

### Using estimates from `tspa()` function

```{r include=FALSE}
# dem60 and ind60
latent_estimates <- data.frame(lavPredict(tspa_fit_2))
lavInspect(tspa_fit_2, what = "est")
# update(tspa_fit_2, data = lavInspect(tspa_fit_2, what = "data"))
# class(tspa_fit_2)

plot(latent_estimates$ind60, 
     latent_estimates$dem60, 
     main = "Scatterplot", 
     xlab = "ind60", 
     ylab = "dem60", 
     pch = 16, 
     ps = 1.25, 
     cex.lab = 1.2, 
     cex.axis = 1.2)

slope <- coef(tspa_fit_2)["dem60~ind60"]
abline(a = 0, b = slope, lwd = 2)
```

## Multigroup, single predictor

```{r}
model <- ' 
  # latent variable definitions
    visual =~ x1 + x2 + x3
    speed =~ x7 + x8 + x9

  # regressions
    visual ~ speed
'

# get factor scores
fs_dat_visual <- get_fs(model = "visual =~ x1 + x2 + x3",
                        data = HolzingerSwineford1939, 
                        group = "school")
fs_dat_speed <- get_fs(model = "speed =~ x7 + x8 + x9", 
                       data = HolzingerSwineford1939, 
                       group = "school")
fs_hs <- cbind(fs_dat_visual, fs_dat_speed)

tspa_fit_3 <- tspa(model = "visual ~ speed",
                   data = fs_hs,
                   se_fs = data.frame(visual = c(0.3391326, 0.311828),
                                      speed = c(0.2786875, 0.2740507)),
                   group = "school"
                   # group.equal = "regressions"
                   )
```

```{r example_multiple_group_single_factor}
tspa_plot(tspa_fit_3,
          ps = 10,
          col = "darkgray",
          cex.lab = 1.2,
          ask = FALSE,
          cex.axis = 1)

tspa_plot(tspa_fit_3,
          ps = 10,
          col = "darkgray",
          cex.lab = 1.2,
          cex.axis = 1,
          ask = TRUE,
          abbreviation = FALSE)
```
